# LUSC vs LUAD

Extraction of expression data from pancancer data set according to cancer type abbreviation:

```{r}
library(dplyr)
LUSC_annot = tcga_annot %>% filter(cancer_type_abbreviation == "LUSC")
LUSC_patients = LUSC_annot$sample
LUSC_exp = tcga_exp_proteins_shortened[, colnames(tcga_exp_proteins_shortened) %in% LUSC_patients, drop = F]

LUAD_annot = tcga_annot %>% filter(cancer_type_abbreviation == "LUAD")
LUAD_patients = LUAD_annot$sample
LUAD_exp = tcga_exp_proteins_shortened[, colnames(tcga_exp_proteins_shortened) %in% LUAD_patients, drop = F]

t_LUAD_exp = as.data.frame(t(LUAD_exp))
t_LUSC_exp = as.data.frame(t(LUSC_exp))

LUSC_LUAD_exp = rbind(t_LUAD_exp, t_LUSC_exp)
```

We extracted and combined LUSC (498 patients) und LUAD (515 patients) expression data into one bigger dataframe containing their expression values.

Now we perform logistic regression based on pathway activity:

First we add a label to distinguish between the two types of tumor. LUAD will be 1 and LUSC will be 0:

```{r}
# gsva_es_comb_ll is a matrix containing enrichment scores based an gsva performed on those two tumor types. 
t_gsva_es_comb_ll = as.data.frame(t(gsva_es_comb_ll))
t_gsva_es_comb_ll$Label <- NA

# Label LUAD
t_gsva_es_comb_ll[rownames(t_gsva_es_comb_ll) %in% LUAD_patients, "Label"] <- 1

# Label LUSC
t_gsva_es_comb_ll[rownames(t_gsva_es_comb_ll) %in% LUSC_patients, "Label"] <- 0
```

Pathways with high variance are stored and selected for the first regression model:

```{r}
variances_gsva <- apply(t(gsva_es_comb_ll), 2, var)
top_variances_gsva <- head(sort(variances_gsva, decreasing = TRUE), 30)
selected_variant_pathways = names(top_variances_gsva)
```

Logistic regression is carried out:

```{r}
# Preparing data: 

# creating a df with selected pathways based on variance

df_selected <- gsva_es_comb_ll[rownames(gsva_es_comb_ll) %in% selected_variant_pathways, ]

t_df_selected = as.data.frame(t(df_selected))


t_df_selected$Label <- NA
# Label LUAD
t_df_selected[rownames(t_df_selected) %in% LUAD_patients, "Label"] <- 1
# Label LUSC
t_df_selected[rownames(t_df_selected) %in% LUSC_patients, "Label"] <- 0


# splitting the df into test and training data
library(caret)
set.seed(123) 
train_index_LUSC_LUAD_2 = createDataPartition(t_df_selected$Label, p = 0.8, list = F) # 60% for training-df
train_data_ll_2 = t_df_selected[train_index_LUSC_LUAD_2, ]
test_data_ll_2 = t_df_selected[-train_index_LUSC_LUAD_2, ]
test_data_model_ll_2 = test_data_ll_2[, !colnames(test_data_ll_2) %in% "Label"]


# logistic regression:

model_ll_2 = glm(Label ~ ., data = train_data_ll_2, family = binomial(link = "logit"))

# making predictions
predictions_ll_2 = predict(model_ll_2, newdata = test_data_model_ll_2, type = "response")

threshold_ll_2 = 0.7
predicted_labels_ll_2 = ifelse(predictions_ll_2 > threshold_ll_2, 1, 0)
accuracy_ll_2 = mean(predicted_labels_ll_2 == test_data_ll_2$Label)
```

Now, backward selection is used to select for pathways to improve the initial model:

```{r}
backwardsmodel <- step(model_ll_2, direction = "backward")
```

```{r}
#HALLMARK_E2F_TARGETS + HALLMARK_G2M_CHECKPOINT + HALLMARK_MITOTIC_SPINDLE + 
    #HALLMARK_MTORC1_SIGNALING + HALLMARK_MYC_TARGETS_V1 + HALLMARK_NOTCH_SIGNALING + 
    #HALLMARK_OXIDATIVE_PHOSPHORYLATION + HALLMARK_PROTEIN_SECRETION + 
   # HALLMARK_UNFOLDED_PROTEIN_RESPONSE + HALLMARK_WNT_BETA_CATENIN_SIGNALING + 
  #  CSR_EARLY_UP.V1_UP + E2F1_UP.V1_UP + EIF4E_DN + RPS14_DN.V1_DN + 
  #  STK33_NOMO_UP + TBK1.DF_DN + TBK1.DN.48HRS_DN + VEGF_A_UP.V1_DN


selected_selected_pathways = c("HALLMARK_E2F_TARGETS", "HALLMARK_G2M_CHECKPOINT", "HALLMARK_MITOTIC_SPINDLE", "HALLMARK_MTORC1_SIGNALING", "HALLMARK_MYC_TARGETS_V1", "HALLMARK_NOTCH_SIGNALING", "HALLMARK_OXIDATIVE_PHOSPHORYLATION", "HALLMARK_PROTEIN_SECRETION", "HALLMARK_UNFOLDED_PROTEIN_RESPONSE", "HALLMARK_WNT_BETA_CATENIN_SIGNALING", "CSR_EARLY_UP.V1_UP", "E2F1_UP.V1_UP", "EIF4E_DN", "RPS14_DN.V1_DN", "STK33_NOMO_UP", "TBK1.DF_DN", "TBK1.DN.48HRS_DN", "VEGF_A_UP.V1_DN")

selected_selected_pathways = c("HALLMARK_MITOTIC_SPINDLE", "HALLMARK_MTORC1_SIGNALING", "HALLMARK_NOTCH_SIGNALING", "HALLMARK_OXIDATIVE_PHOSPHORYLATION", "HALLMARK_PROTEIN_SECRETION", "HALLMARK_WNT_BETA_CATENIN_SIGNALING", "E2F1_UP.V1_UP", "EIF4E_DN", "RPS14_DN.V1_DN", "STK33_NOMO_UP", "TBK1.DF_DN", "TBK1.DN.48HRS_DN", "VEGF_A_UP.V1_DN")
```

The final logistic regression model:

```{r}
df_selected_selected <- gsva_es_comb_ll[rownames(gsva_es_comb_ll) %in% selected_selected_pathways, ]

t_df_selected_selected = as.data.frame(t(df_selected_selected))


t_df_selected_selected$Label <- NA

# Label LUAD
t_df_selected_selected[rownames(t_df_selected_selected) %in% LUAD_patients, "Label"] <- 1
# Label LUSC
t_df_selected_selected[rownames(t_df_selected_selected) %in% LUSC_patients, "Label"] <- 0


# splitting the df into test and training data
set.seed(123) 
train_index_LUSC_LUAD_4 = createDataPartition(t_df_selected_selected$Label, p = 0.8, list = F) # 80% for training-df
train_data_ll_4 = t_df_selected_selected[train_index_LUSC_LUAD_4, ]
test_data_ll_4 = t_df_selected_selected[-train_index_LUSC_LUAD_4, ]
test_data_model_ll_4 = test_data_ll_4[, !colnames(test_data_ll_4) %in% "Label"]


# logistic regression
model_ll_4 = glm(Label ~ ., data = train_data_ll_4, family = binomial(link = "logit"))

# making predictions
predictions_ll_4 = predict(model_ll_4, newdata = test_data_model_ll_4, type = "response")

threshold_ll_4 = 0.5
predicted_labels_ll_4 = ifelse(predictions_ll_4 > threshold_ll_4, 1, 0)
accuracy_ll_4 = mean(predicted_labels_ll_4 == test_data_ll_4$Label)
```

Here, a confusion matrix is created:

```{r}
confusion_matrix_LL <- table(predicted_labels_ll_4, test_data_ll_4$Label)
confusion_matrix_LL
```
plotting confusion matrix
```{r}
#install.packages("cvms")

data_eval = data.frame(
  "predictions" = predicted_labels_ll_4, 
  "target" = test_data_ll_4$Label
)

library(cvms)
eval <- evaluate(
  data = data_eval,
  target_col = "target",
  prediction_cols = "predictions",
  type = "binomial"
)


plot_confusion_matrix(eval)

plot_confusion_matrix(eval$`Confusion Matrix`[[1]],
                      add_sums = T)
```



```{r}
# Extraction of relevant values from the confusion matrix 
TN <- confusion_matrix_LL[1, 1]  # True Negatives
FP <- confusion_matrix_LL[2, 1]  # False Positives
FN <- confusion_matrix_LL[1, 2]  # False Negatives
TP <- confusion_matrix_LL[2, 2]  # True Positives

accuracy_ll_4 <- (TP + TN) / (TP + TN + FP + FN)  # Accuracy
FNR <- FN / (FN + TP)  # False-Negative Rate
FPR <- FP / (FP + TN)  # False-Positive Rate
TPR <- TP / (TP + FN)  # True-Positive Rate (Recall)

# printing results
cat("Accuracy: ", accuracy_ll_4, "\n")
cat("False-Negative Rate: ", FNR, "\n")
cat("False-Positive Rate: ", FPR, "\n")
cat("True-Positive Rate (Recall): ", TPR, "\n")
```

# Statistical tests, to evaluate the model

Waldtest:
```{r}
library(lmtest)
wald_test <- summary(model_ll_4)$coefficients[, c("Estimate", "Std. Error", "z value", "Pr(>|z|)")]
wald_test
```
All coefficients are significant (p < 0.05).


McFadden's pseudo R^2 test:
```{r}
logLik_full <- logLik(model_ll_4)

# Null-Modell anpassen und Log-Likelihood berechnen
null_model <- glm(Label ~ 1, data = train_data_ll_4, family = "binomial")
logLik_null <- logLik(null_model)

# McFadden's pseudo R^2 berechnen
McFadden_R2 <- 1 - (logLik_full / logLik_null)
McFadden_R2
```
McFadden’s pseudo-R2 = 0.7075055 meaning that the model is able to explain 70.75% of the variance. 


Deviance test:
```{r}
deviance_test <- anova(null_model, model_ll_4, test = "LRT")
deviance_test
```


Chi-squared test:
```{r}
chi_sq_test <- chisq.test(confusion_matrix_LL)
chi_sq_test
```
The values 2.2e-16 indicates, that the model is significantly better than a random model in predicting the tumor type. 


# ROC-curve

```{r}
#install.packages("ISLR")
library(ISLR)
library(ROCR)

# Load a binary classification dataset from ISLR package
t_df_selected_selected <- ISLR::Default
# Plot ROC curve
pred  <- prediction(predictions_ll_4, test_data_ll_4$Label)
roc_curve <- performance(pred, "tpr", "fpr")

perf <- performance(pred,"auc")
auc <- as.numeric(perf@y.values)

#plot(roc_curve, col = "dodgerblue3", main = "ROC Curve", lwd = 2)
#abline(0, 1, col = "black", lty = 2, lwd = 1.5)
#text(0.5, 0.3, paste("AUC = ", round(auc, 4)), adj = c(0.5, 0.5), col = "black", cex = 0.9)

library(ggplot2)
roc_df <- data.frame(
  FPR = roc_curve@x.values[[1]],
  TPR = roc_curve@y.values[[1]]
)

# Plot mit ggplot2 erstellen
ggplot(roc_df, aes(x = FPR, y = TPR)) +
  geom_line(color = "dodgerblue3", size = 1.2) +
  geom_abline(slope = 1, intercept = 0, linetype = "dashed", color = "black", size = 0.7) +
  annotate("text", x = 0.5, y = 0.3, label = paste("AUC =", round(auc, 4)), color = "black", size = 3.8) +
  labs(title = "ROC Curve", x = "False Positive Rate", y = "True Positive Rate") +
  theme_minimal()
```

# Appendix with further figures

Plot of distribution of LUAD and LUSC patients after PCA:

```{r}
pca_LUSC_LUAD = prcomp(t(df_selected_selected))

colors = c(rep("LUAD", 515), rep("LUSC", 498))
# Erstellen des UMAP-Plots
ggplot(pca_LUSC_LUAD$x, aes(x = PC1, y = PC2, color = colors)) +
  geom_point(size = 2) 
```

UMAP plot: LUAD vs LUSC 
a) without PCA 
```{r}
library(umap)
library(ggplot2)

# transpose matrix
gsva_es_comb_ll_t <- t(gsva_es_comb_ll)
```

```{r}
# calculate umap
umap_gsva_ll <- umap(gsva_es_comb_ll_t)

# check dimension
print(dim(umap_gsva_ll$layout))  # should be n x 2 
```

```{r}
plot(umap_gsva_ll$layout[,1], umap_gsva_ll$layout[,2],
     xlab = "UMAP Dimension 1", ylab = "UMAP Dimension 2",
     main = "UMAP Plot",
     pch = 20)
```

```{r}
# annotations 
annot <- rbind(LUSC_annot, LUAD_annot)

# combine data
umap_gsva_ll_color <- data.frame(
  UMAP1 = umap_gsva_ll$layout[, 1],
  UMAP2 = umap_gsva_ll$layout[, 2],
  tumor = annot$cancer_type_abbreviation)

# plot umap results
ggplot(umap_gsva_ll_color, aes(x = UMAP1, y = UMAP2, color = tumor)) +
  geom_point(size = 1) +
  labs(title = "UMAP plot: LUAD vs LUSC",
       x = "UMAP dimension 1",
       y = "UMAP dimension 2") +
  theme_minimal() +
  theme(legend.position = "right")  

# define color 
custom_colors <- c("LUAD" = "dodgerblue", "LUSC" = "brown1")

# plot with own colors
ggplot(umap_gsva_ll_color, aes(x = UMAP1, y = UMAP2, color = tumor)) +
  geom_point(size = 1) +
  labs(title = "UMAP Plot: LUAD vs LUSC",
       x = "UMAP Dimension 1",
       y = "UMAP Dimension 2") +
  scale_color_manual(values = custom_colors, name = "Tumor Type") +
  theme_minimal() +
  theme(legend.position = "right")

```

b) umap computed on pca results
```{r}
# PCA 
library(irlba)
pca_gsva_ll <- prcomp(gsva_es_comb_ll_t)

```

```{r}
scores_ll <- pca_gsva_ll$x
```

```{r}
# calculate umap
umap_pca_gsva_ll <- umap(scores_ll)

# check dimension
print(dim(umap_pca_gsva_ll$layout))  # should be n x 2 
```

```{r}
# combine data
umap_pca_gsva_ll_color <- data.frame(
  UMAP1 = umap_pca_gsva_ll$layout[, 1],
  UMAP2 = umap_pca_gsva_ll$layout[, 2],
  tumor = annot$cancer_type_abbreviation)

# plot umap results
ggplot(umap_pca_gsva_ll_color, aes(x = UMAP1, y = UMAP2, color = tumor)) +
  geom_point(size = 1) +
  labs(title = "UMAP plot: LUAD vs LUSC",
       x = "UMAP dimension 1",
       y = "UMAP dimension 2") +
  theme_minimal() +
  theme(legend.position = "right")  
```

Plot
```{r}
data_reg_plot = data.frame(
  "HALLMARK_WNT_BETA_CATENIN_SIGNALING" = t_df_selected_selected$HALLMARK_WNT_BETA_CATENIN_SIGNALING,
  "Label" = t_df_selected_selected$Label)

model_reg_plot <- glm(Label ~ HALLMARK_WNT_BETA_CATENIN_SIGNALING, data = data_reg_plot, family = binomial)


new_data <- data.frame(HALLMARK_WNT_BETA_CATENIN_SIGNALING = seq(min(t_df_selected_selected$HALLMARK_WNT_BETA_CATENIN_SIGNALING), max(t_df_selected_selected$HALLMARK_WNT_BETA_CATENIN_SIGNALING), length.out = length(t_df_selected_selected$HALLMARK_WNT_BETA_CATENIN_SIGNALING)))

new_data$prob <- predict(model_reg_plot, newdata = new_data, type = "response")


library(ggplot2)

ggplot(data_reg_plot, aes(x = HALLMARK_WNT_BETA_CATENIN_SIGNALING, y = Label)) +
  geom_point(alpha = 0.3, position = position_jitter(width = 0.1, height = 0.05)) +
  geom_line(data = new_data, aes(x = HALLMARK_WNT_BETA_CATENIN_SIGNALING, y = prob), color = "blue", size = 1.2) +
  labs(title = "Logistic Regression Model", x = "HALLMARK_WNT_BETA_CATENIN_SIGNALING activity", y = "Probability of being LUAD") +
  theme_minimal() +
  theme(plot.title = element_text(hjust = 0.5))
```
```{r}
data_reg_plot_2 = data.frame(
  "VEGF_A_UP.V1_DN" = t_df_selected_selected$VEGF_A_UP.V1_DN,
  "Label" = t_df_selected_selected$Label)

model_reg_plot_2 <- glm(Label ~ VEGF_A_UP.V1_DN, data = data_reg_plot_2, family = binomial)


new_data_2 <- data.frame(VEGF_A_UP.V1_DN = seq(min(t_df_selected_selected$VEGF_A_UP.V1_DN), max(t_df_selected_selected$VEGF_A_UP.V1_DN), length.out = length(t_df_selected_selected$VEGF_A_UP.V1_DN)))

new_data_2$prob <- predict(model_reg_plot_2, newdata = new_data_2, type = "response")


library(ggplot2)

ggplot(data_reg_plot_2, aes(x = VEGF_A_UP.V1_DN, y = Label)) +
  geom_point(alpha = 0.3, position = position_jitter(width = 0.1, height = 0.05)) +
  geom_line(data = new_data_2, aes(x = VEGF_A_UP.V1_DN, y = prob), color = "blue", size = 1.2) +
  labs(title = "Logistic Regression Model", x = "VEGF_A_UP.V1_DN activity", y = "Probability of being LUAD") +
  theme_minimal() +
  theme(plot.title = element_text(hjust = 0.5))