---
title: ''
output:
  pdf_document:
    number_sections: yes
  html_document:
    code_folding: hide
  word_document: default
documentclass: scrreprt
classoption:
- parskip
- oneside
geometry:
- left=2.00cm
- right=2.00cm
- top=1.50cm
- bottom=1.50cm
- bindingoffset=0.1cm
- includehead
- includefoot
linestretch: 1.15
header-includes:
- \usepackage[greek, ngerman, main=english]{babel}
- \usepackage[utf8]{inputenc}
- \usepackage[T1]{fontenc}
- \usepackage{lmodern}
- "\\usepackage[headsepline]{scrlayer-scrpage}"
- \usepackage{url}
- \usepackage[backend=biber, style=authoryear, giveninits=true, maxbibnames=99, uniquename=init,
  maxcitenames=2, hyperref=true, date=year]{biblatex}
  \DeclareUnicodeCharacter{2061}{}
  \DeclareUnicodeCharacter{2032}{}
  \DeclareUnicodeCharacter{2223}{}
- \usepackage{xpatch}
- \usepackage{csquotes}
- \usepackage{amsmath}
- \usepackage{listings}
- \usepackage{booktabs}
- \usepackage{longtable}
- \usepackage{multirow}
- \usepackage{rotating}
- \usepackage{subfigure}
- \usepackage{graphicx}
- \usepackage{float}
- \usepackage{acronym}
- \usepackage{lipsum}
- \usepackage{scrhack}
- \usepackage{etoolbox}
- \makeatletter
- \patchcmd{\scr@startchapter}{\if@openright\cleardoublepage\else\clearpage\fi}{}{}{}
- \makeatother
- \emergencystretch=50pt
- \clubpenalty = 10000
- \widowpenalty = 10000
- \displaywidowpenalty = 10000
- \automark[section]{chapter}
- \renewcommand*{\chaptermarkformat}{}
- \renewcommand*{\sectionmarkformat}{}
- \renewcommand*{\chapterheadstartvskip}{\vspace*{0.1cm}}
- \renewcommand*{\chapterheadendvskip}{\vspace*{0.1cm}}
- \setkomafont{title}{\sffamily}
- \setkomafont{disposition}{\usekomafont{title}}
- \setkomafont{author}{\usekomafont{title}}
- \setkomafont{date}{\usekomafont{title}}
- \setkomafont{caption}{\sffamily\small}
- \setkomafont{captionlabel}{\usekomafont{caption}\bfseries\small}
- \setkomafont{pagehead}{\normalfont\scshape}
csl: "bib/molecular-biology-of-the-cell.csl"
bibliography:
- bib/packages.bib
- bib/test.bib
---
<!-- Based on Nicolas Peschkes LaTeX Template for MoBis, if you want further information on the packages just check it here: https://gitlab.com/npeschke/LaTeX-Template-IPMB -->

<!-- your title page -->


\begin{titlepage}
\centering
	{\Large Ruprecht-Karls-Universität Heidelberg\\
		Fakultät für Ingenieurswissenschaften\\
		Bachelorstudiengang Molekulare Biotechnologie\\}

	{\vspace{\stretch{2}}}
	{\usekomafont{title}

		{\Huge Comparative Pan-Cancer Analysis Reveals Pathway Signatures Distinguishing Lung Adenocarcinoma from Squamous Cell Carcinoma}

		{\Huge}

		{\Huge}

	}

	\vspace{\stretch{2}}
	{\Large Data Science Project SoSe 2024}

	\vspace{\stretch{2}}

	{\normalsize
		\begin{tabular}{rl}
			Autoren: Elisabeth Werner, Julia Grigorjew, Christopher Kolb, Paul Averdung\\
			Heidelberg, 08.07.2024\\
			Supervisor: Prof. Dr. Carl Herrmann\\
		\end{tabular}
	}

	\vspace{\stretch{1}}

\end{titlepage}

<!-- Abkürzungverzeichnis -->

\tableofcontents

\clearpage
<!-- Länge des Strings in den [..] entspricht dem Abstand der zwischen Acronym und Vollname gelassen wird -->
<!-- 
\acro{acronym}{full name}
\acused{acronym}
-->
<!-- add your chapters here -->

```{r child = "chapter/abstract.Rmd"}
```
\addchap{Abbreviations}

\begin{acronym}[XXXX]
\acro{ACC}{Adrenocortical carcinoma}
\acro{AIC}{Akaike Information Criterion}
\acro{AUC}{Area under the curve}
\acro{BLCA}{Bladder urothelial carcinoma}
\acro{BRCA}{Breast invasive carcinoma}
\acro{CESC}{Cervical squamous cell carcinoma and endocervical adenocarcinoma}
\acro{CHOL}{Cholangiocarcinoma}
\acro{COAD}{Colon adenocarcinoma}
\acro{DLBC}{Lymphoid neoplasm diffuse large b-cell lymphoma}
\acro{ES}{Enrichment score}
\acro{ESCA}{Esophageal carcinoma}
\acro{FC}{Fold change}
\acro{GBM}{Glioblastoma multiforme}
\acro{GSEA}{Gene set enrichment analysis}
\acro{GSVA}{Gene set variation analysis}
\acro{KICH}{Kidney Chromophobe}
\acro{KIRC}{Kidney renal clear cell carcinoma}
\acro{KIRP}{Kidney renal papillary cell carcinoma}
\acro{LAML}{Acute myeloid leukemia}
\acro{LGG}{brain lower grade glioma}
\acro{LIHC}{Liver hepatocellular carcinoma}
\acro{LUAD}{Lung adenocarcinoma}
\acro{LUSC}{Lung squamous cell carcinoma}
\acro{MESO}{Mesothelioma}
\acro{MPC}{model predictive control}
\acro{MSigDB}{Molecular signatures database}
\acro{NES}{Normalized enrichment score}
\acro{OV}{Ovarian serous cystadenocarcinoma}
\acro{PAAD}{Pancreatic adenocarcinom}
\acro{PCA}{Principal component analysis}
\acro{PCPG}{Pheochromocytoma and Paraganglioma}
\acro{PRAD}{Prostate adenocarcinoma}
\acro{READ}{Rectum adenocarcinoma}
\acro{ROC}{Receiver operator characteristic}
\acro{SARC}{Sarcoma}
\acro{SKCM}{Skin cutaneous melanoma}
\acro{STAD}{Stomach adenocarcinoma}
\acro{TCGA}{The Cancer Genome Atlas}
\acro{TGCT}{Testicular germ cell tumors}
\acro{THCA}{Thyroid carcinoma}
\acro{THYM}{Thymoma}
\acro{TPM}{Transcripts per million}
\acro{UCEC}{Uterine corpus endometrial carcinoma}
\acro{UCS}{Uterine carcinosarcoma}
\acro{UMAP}{Uniform manifold approximation and projection}
\acro{UVM}{Uveal melanoma}
\end{acronym}
\newpage
```{r child = "chapter/introduction.Rmd"}
```

```{r child = "chapter/material_and_methods.Rmd"}
```

```{r child = "chapter/results.Rmd"}
```

```{r child = "chapter/discussion.Rmd"}
```

```{r child = "chapter/appendix.Rmd"}
```
\addchap{References}
\printbibliography