Installation of the msigdbr package -> for data sets:
```{r}
install.packages("msigdbr")
```
Load Library:
```{r}
library(msigdbr)
```
Load all gene sets of human genome:
```{r}
all_gene_sets = msigdbr(species = "Homo sapiens")
```
Show different gene set collections of the package:
```{r}
msigdbr_collections()
```
Show hallmark genes for human:
```{r}
hallmark_gene_sets = msigdbr(species = "human", category = "H")
head(hallmark_gene_sets)
```
Gene sets hallmarks interesting:
1. HALLMARK_EPITHELIAL_MESENCHYMAL_TRANSITION: This gene set is associated with the transition of epithelial cells to a mesenchymal state, which is a critical process in cancer metastasis, including lung adenocarcinoma.
2. HALLMARK_APOPTOSIS: Apoptosis, or programmed cell death, is a fundamental process that is often dysregulated in cancer, including lung adenocarcinoma. Analyzing this gene set can provide insights into apoptotic pathways involved in cancer progression and therapy response.
3. HALLMARK_ANGIOGENESIS: Angiogenesis, the formation of new blood vessels, is a hallmark of cancer and is crucial for tumor growth and metastasis. Analyzing this gene set can shed light on angiogenic processes that contribute to lung adenocarcinoma progression.
4. HALLMARK_E2F_TARGETS: The E2F transcription factor targets gene set represents genes regulated by E2F transcription factors, which play key roles in cell cycle regulation and proliferation. Dysregulation of E2F targets is common in cancer, including lung adenocarcinoma.
5. HALLMARK_MYC_TARGETS_V1: MYC is a well-known oncogene that regulates genes involved in cell growth, proliferation, and metabolism. Analyzing this gene set can provide insights into MYC-regulated pathways that are relevant to lung adenocarcinoma.
```{r}
hallmark_gene_set_epithelial_mesenchymal_transition <- hallmark_gene_sets[2122:2325,]
hallmark_gene_set_apoptosis <- hallmark_gene_sets[961:1143,]
hallmark_gene_set_angiogenesis <- hallmark_gene_sets[648:683,]
hallmark_gene_set_E2F_targets <- hallmark_gene_sets[1904:2121,]
hallmark_gene_set_MYC_targets_V1 <- hallmark_gene_sets[5642:5937,]
#as row 5687 is the same as row 5686 with another id (which does not match the general pattern of ENSG), I will remove row 46 in the matrix it
hallmark_gene_set_MYC_targets_V1_2 = hallmark_gene_set_MYC_targets_V1[-46,]
```
Next the corresponding emsembl_ids will be stored in a variable to calculate the jaccards index:
```{r}
ensemble_ids_epithelial_mesenchymal_transition = hallmark_gene_set_epithelial_mesenchymal_transition$ensembl_gene
ensemble_ids_apoptosis = hallmark_gene_set_apoptosis$ensembl_gene
ensemble_ids_angiogenesis = hallmark_gene_set_angiogenesis$ensembl_gene
ensemble_ids_E2F_targets = hallmark_gene_set_E2F_targets$ensembl_gene
# for MYC I will use the ids stored in the updated matrix where row 46 is deleted
ensemble_ids_MYC_targets_V1 = hallmark_gene_set_MYC_targets_V1_2$ensembl_gene
```
Now the Jaccards index can be computed (Jaccard Similarity = (number of observations in both sets) / (number in either set)):
```{r}
jaccard <- function(a, b) {
    intersection = length(intersect(a, b))
    union = length(union(a,b))
    return (intersection/union)
}
```

```{r}
# print(ensemble_ids_MYC_targets_V1)
length(intersect(ensemble_ids_MYC_targets_V1, ensemble_ids_MYC_targets_V1)) / length(union(ensemble_ids_MYC_targets_V1, ensemble_ids_MYC_targets_V1))
```
Jaccard for our gene sets:
```{r}
#create list from all the gene set ids 
gene_sets_ids = list(ensemble_ids_angiogenesis, ensemble_ids_apoptosis, ensemble_ids_E2F_targets, ensemble_ids_epithelial_mesenchymal_transition, ensemble_ids_MYC_targets_V1)

#create similarity matrix 
similarity_matrix = matrix(0, nrow =length(gene_sets_ids), ncol = length(gene_sets_ids))
rownames(similarity_matrix) = colnames(similarity_matrix) = names(gene_sets_ids)
for (i in 1:length(gene_sets_ids)) {
  for (j in 1:length(gene_sets_ids)) {
    similarity_matrix[i, j] = jaccard(gene_sets_ids[[i]], gene_sets_ids[[j]])
  }
}
```
Plotting the result of the jaccard similarities in a heatmap:
```{r}
names = c("angiogenesis", "apoptosis", "E2F_targets", "ep_mes_trans", "MYC_targets")
library(pheatmap)
pheatmap(as.matrix(similarity_matrix),
          Rowv = NA, Colv = NA,
         cluster_rows = F,
         cluster_cols = F,
          scale = 'none',
          key = T,
         labels_row = names,
         labels_col = names,
          main = "Jaccard Similarity Heatmap")
# low similarity amongs the different data sets can be observed
```
Finally the Jaccard Index of the protein encoding genes from the tcga data set and the gene sets needs to be computed:
```{r}
#ensemble ids of proteins in tcga data set
ensemble_ids_tcga_proteins = tcga_exp_proteins_shortened$ensembl_gene_id
# compute jaccard index of tcga proteins and gene sets 
jaccard(ensemble_ids_tcga_proteins, ensemble_ids_tcga_proteins)
jaccard(ensemble_ids_tcga_proteins,ensemble_ids_epithelial_mesenchymal_transition)
jaccard(ensemble_ids_tcga_proteins,ensemble_ids_apoptosis)
jaccard(ensemble_ids_tcga_proteins,ensemble_ids_angiogenesis)
jaccard(ensemble_ids_tcga_proteins,ensemble_ids_E2F_targets)
jaccard(ensemble_ids_tcga_proteins,ensemble_ids_MYC_targets_V1)
```

Vergleich zu tcga_exp_proteins --> alle Proteincodierende Gene, auch mit niedriger Varianz
```{r}
#ensemble ids of proteins in tcga data set
ensemble_ids_tcga_proteins = tcga_exp_filtered_proteins$ensembl_gene_id
# compute jaccard index of tcga proteins and gene sets 
jaccard(ensemble_ids_tcga_proteins, ensemble_ids_tcga_proteins)
jaccard(ensemble_ids_tcga_proteins,ensemble_ids_epithelial_mesenchymal_transition)
jaccard(ensemble_ids_tcga_proteins,ensemble_ids_apoptosis)
jaccard(ensemble_ids_tcga_proteins,ensemble_ids_angiogenesis)
jaccard(ensemble_ids_tcga_proteins,ensemble_ids_E2F_targets)
jaccard(ensemble_ids_tcga_proteins,ensemble_ids_MYC_targets_V1)
```

