Here we want to select and analyse suitable gene sets from msigdb based on our previous analysis
```{r}
BiocManager::install("fgsea")
install.packages("msigdbr")
library(msigdbr)
library(tidyverse)
library(clusterProfiler)
library(RColorBrewer)
library(fgsea)
```
As can be seen by the GSEA Analysis we performed over the LUAD Data with the hallmark gene sets provides by Hermann we can see that Repair-, Apoptosis-, Metastasis-, Proliferation- and Replication Pathways are especially significant enriched in LUAD. Thus we will focus on Genes belonging to these areas. 

So I decided to take a look at the gene ontology gene sets (C5) (subcategory = biological processes) from msigdb
Please copy code to you DATA folder and remove #s there in order to run code as it is to big to be pushed
```{r}
#library("dplyr")
#C5 = msigdbr(species = "Homo sapiens", category = "C5", subcategory = "GO:BP")
#str(C5)
#class(hallmark_gene_sets)
#we need to reformat the msigdbr output into a list of vectors (gene sets) with the corresponding genes in them

#very important code for GSEA with msigdb
#C5.genesymbols.ls = C5 %>%
  #dplyr::select(gs_name, gene_symbol) %>%
  #group_by(gs_name) %>%
  #summarise(all.genes = list(unique(gene_symbol))) %>%
  #deframe()
```
C6 optimized for GSEA
```{r}
#library("dplyr")
#C6 = msigdbr(species = "Homo sapiens", category = "C6")
#str(C6)
#class(hallmark_gene_sets)
#we need to reformat the msigdbr output into a list of vectors (gene sets) with the corresponding genes in them

#very important code for GSEA with msigdb
#C6.genesymbols.ls = C6 %>%
  #dplyr::select(gs_name, gene_symbol) %>%
  #group_by(gs_name) %>%
  #summarise(all.genes = list(unique(gene_symbol))) %>%
  #deframe()
```
These can now be analysed the same way we did with the other gene sets
```{r}
similarity_matrix_msigdbC5 = matrix(0, nrow = length(C5.genesymbols.ls), ncol = length(C5.genesymbols.ls))
rownames(similarity_matrix_msigdbC5) = colnames(similarity_matrix_msigdbC5) = names(C5.genesymbols.ls)
for (i in 1:length(C5.genesymbols.ls)) {
  for (j in 1:length(C5.genesymbols.ls)) {
    similarity_matrix_msigdbC5[i, j] = jaccard(C5.genesymbols.ls[[i]], C5.genesymbols.ls[[j]])
  }
}
```
Code above didnt work as to much time was needed. Thus I will try the same thing with the C6 oncogenic signature genes:
```{r}
similarity_matrix_msigdbC6 = matrix(0, nrow = length(C6.genesymbols.ls), ncol = length(C6.genesymbols.ls))
rownames(similarity_matrix_msigdbC6) = colnames(similarity_matrix_msigdbC6) = names(C6.genesymbols.ls)
for (i in 1:length(C6.genesymbols.ls)) {
  for (j in 1:length(C6.genesymbols.ls)) {
    similarity_matrix_msigdbC6[i, j] = jaccard(C6.genesymbols.ls[[i]], C6.genesymbols.ls[[j]])
  }
}
```
Now we will plot the similarities in a heatmap
```{r}
names = names(C6.genesymbols.ls)
library(pheatmap)
pheatmap(as.matrix(similarity_matrix_msigdbC6),
      cellwidth = 5, cellheight = 5,
        cluster_rows = F,
         cluster_cols = F,
         fontsize = 5,
          main = "Jaccard Similarity Heatmap")
```
We can tell from the heatmap above that there is no significant overlap in any of the gene sets. Therefore we now perform a GSEA of the LUAD data with these gene sets
```{r}
fgsea_res_msigdbC6 = fgseaSimple(pathways = C6.genesymbols.ls,
                         stats = fold_change_ranked_list_gn,
                         nperm = 10000,
                         minSize = 1,
                         maxSize = length(fold_change_ranked_list_gn) - 1)
```

```{r}
# Plot GSEA results using barplot
barplot(fgsea_res_msigdbC6$NES, 
        names.arg = fgsea_res_msigdbC6$pathway, 
        las = 2,
        cex.names = 0.3,
        col = ifelse(fgsea_res_msigdbC6$padj < 0.05, "red", "gray"),
        main = "GSEA Results C6",
        xlab = "Pathway",
        ylab = "Normalized Enrichment Score (NES)",
)
```
Then we will select those gene sets that showed significant enrichment (23 gene sets out of the 189)
```{r}
C6_significant_gs = fgsea_res_msigdbC6$pathway[which(fgsea_res_msigdbC6$padj < 0.05)]
print(C6_significant_gs)
C6.new.genesets = C6.genesymbols.ls[C6_significant_gs]
C6.all.genesets = C6.genesymbols.ls
```
