At first we try to convert the gene names in the expression data and in the gene sets we use to ensemble IDs

So we rank the fold changes of the patients and link the values to the ensemble IDs
```{r}
rank_patients_ensemble = list()

for (patient in colnames(fold_changes_patients)){
  patient_fc_ens = fold_changes_patients[, patient]
  names(patient_fc_ens) = rownames(fold_changes_patients)
  rank_patients_ensemble[[patient]] = (sort(patient_fc_ens, decreasing = T))
}
library(fgsea)
library(pheatmap)
```
Next we convert the genesets we have to ensemble IDs (nix gut diese)
```{r}
ensembl = useEnsembl(biomart = "genes", dataset = "hsapiens_gene_ensembl")
ensembl_lookup = getBM(attributes = c("ensembl_gene_id", "external_gene_name"),
                       mart = ensembl)
new_genesets_ens1 = sapply(1:length(new_genesets), function(x) {
  ensembl_lookup$ensembl_gene_id[which(ensembl_lookup$external_gene_name %in% new_genesets[[x]])]
})
names(new_genesets_ens1) = names(new_genesets)
```
We now need to create new named numericals from the list, as type list is not accepted by fgsea
```{r}
named.numericals.patients.ens = list()
for(patient in names(rank_patients_ensemble)){
 named.num.ens = unlist(rank_patients_ensemble[patient])
 names(named.num.ens) = names(rank_patients_ensemble[[patient]])
 named.numericals.patients.ens[[patient]] = named.num.ens
}
```
Next we compute the GSEA on the patients
```{r}
NES.list.patients.ens = list()
for (patient in names(named.numericals.patients.ens)){
  stats.gsea.patients.ens = named.numericals.patients.ens[[patient]]
  gsea_patient_res.ens = fgseaSimple(
    stats = stats.gsea.patients.ens,
    nperm = 100,
    pathways = new_genesets_ens1
  )
  NES.list.patients.ens[[patient]] = gsea_patient_res.ens$NES
}
```
Plotting the NES in a heat map 
```{r}
enrichment_matrix.ens = as.matrix(NES.list.patients.ens[[1]])
for (i in (2:58)){
  vec.ens = as.matrix(NES.list.patients.ens[[i]])
  enrichment_matrix.ens = cbind(enrichment_matrix.ens, vec.ens)
}
rownames(enrichment_matrix.ens) = names(new_genesets)
colnames(enrichment_matrix.ens) = numbers_patients

heatm.patietns.gsea.ens = pheatmap(enrichment_matrix.ens, cluster_rows = , labels_row = names_new_genesets, fontsize_row = 5, labels_col = numbers_patients, fontsize = 7) 
```
```{r}
#saveRDS(enrichment_matrix.ens, "/Users/paulaverdung/Desktop/Uni/Bachelor MoBi/Info/Bioinfoprojekt/NES.enrichment_matrix.ens.RDS")
```
Next compute matrix with insignificant NES (p val) set to 0 
```{r}
pval.list.patients.ens = list()
for (patient in names(named.numericals.patients.ens)){
  stats.gsea.patients.ens = named.numericals.patients.ens[[patient]]
  gsea_patient_res.ens = fgseaSimple(
    stats = stats.gsea.patients.ens,
    nperm = 100,
    pathways = new_genesets_ens1
  )
  pval.list.patients.ens[[patient]] = gsea_patient_res.ens$pval
}
```
```{r}
#store the adjusted p vals in a matrix of same format as the NES matrix
pval_matrix.ens = as.matrix(pval.list.patients.ens[[1]])
for (i in (2:58)){
  p.vec.ens = as.matrix(pval.list.patients.ens[[i]])
  pval_matrix.ens = cbind(pval_matrix.ens, p.vec.ens)
}
#find positions in which p-values are above certain threshhold
pval.t.ens = 0.15
nsig.NES.ens = which(pval_matrix.ens > pval.t.ens)
adj.enrichment_matrix.ens = enrichment_matrix.ens
for (e in nsig.NES.ens){
  adj.enrichment_matrix.ens[e] = 0
}
```
Now plot new heatmap
```{r}
adj.heatm.patietns.gsea.ens = pheatmap(adj.enrichment_matrix.ens, cluster_rows = , labels_row = names_new_genesets, fontsize_row = 5, labels_col = numbers_patients, fontsize = 7)
```
```{r}
#saveRDS(adj.enrichment_matrix.ens, "/Users/paulaverdung/Desktop/Uni/Bachelor MoBi/Info/Bioinfoprojekt/NES.enrichment_matrix.adj.ens.RDS")
```
Same procedere is required for the new genesets C6
Convert to ensembl IDs
```{r}
new_genesets_ens1.C6 = sapply(1:length(C6.new.genesets), function(x) {
  ensembl_lookup$ensembl_gene_id[which(ensembl_lookup$external_gene_name %in% C6.new.genesets[[x]])]
})
names(new_genesets_ens1.C6) = names(C6.new.genesets)
```
Compute GSEA:
```{r}
NES.list.patients.ens.C6 = list()
for (patient in names(named.numericals.patients.ens)){
  stats.gsea.patients.ens.C6 = named.numericals.patients.ens[[patient]]
  gsea_patient_res.ens.C6 = fgseaSimple(
    stats = stats.gsea.patients.ens.C6,
    nperm = 100,
    pathways = new_genesets_ens1.C6
  )
  NES.list.patients.ens.C6[[patient]] = gsea_patient_res.ens.C6$NES
}
```
Plot heatmap unprocessed:
```{r}
enrichment_matrix.ens.C6 = as.matrix(NES.list.patients.ens.C6[[1]])
for (i in (2:58)){
  vec.ens.C6 = as.matrix(NES.list.patients.ens.C6[[i]])
  enrichment_matrix.ens.C6 = cbind(enrichment_matrix.ens.C6, vec.ens.C6)
}
rownames(enrichment_matrix.ens.C6) = names(C6.new.genesets)
colnames(enrichment_matrix.ens.C6) = numbers_patients

heatm.patietns.gsea.ens.C6 = pheatmap(enrichment_matrix.ens.C6, cluster_rows = , labels_row = names(C6.new.genesets), fontsize_row = 5, labels_col = numbers_patients, fontsize = 7) 
````
```{r}
#saveRDS(enrichment_matrix.ens.C6, "/Users/paulaverdung/Desktop/Uni/Bachelor MoBi/Info/Bioinfoprojekt/NES.enrichment_matrix.C6.ens.RDS")
```
Now process the heatmap similarly to above:
```{r}
pval.list.patients.ens.C6 = list()
for (patient in names(named.numericals.patients.ens)){
  stats.gsea.patients.ens.C6 = named.numericals.patients.ens[[patient]]
  gsea_patient_res.ens.C6 = fgseaSimple(
    stats = stats.gsea.patients.ens.C6,
    nperm = 100,
    pathways = new_genesets_ens1.C6
  )
  pval.list.patients.ens.C6[[patient]] = gsea_patient_res.ens.C6$pval
}
```
```{r}
#store the adjusted p vals in a matrix of same format as the NES matrix
pval_matrix.ens.C6 = as.matrix(pval.list.patients.ens.C6[[1]])
for (i in (2:58)){
  p.vec.ens.C6 = as.matrix(pval.list.patients.ens.C6[[i]])
  pval_matrix.ens.C6 = cbind(pval_matrix.ens.C6, p.vec.ens.C6)
}
#find positions in which p-values are above certain threshhold
pval.t.ens.C6 = 0.15
nsig.NES.ens.C6 = which(pval_matrix.ens.C6 > pval.t.ens.C6)
NES.enrichment_matrix.C6.adj.ens = enrichment_matrix.ens.C6
for (e in nsig.NES.ens.C6){
  NES.enrichment_matrix.C6.adj.ens[e] = 0
}
```
Plot the adjusted heatmap:
```{r}
adj.heatm.patietns.gsea.ens.C6 = pheatmap(adj.enrichment_matrix.ens.C6, cluster_rows = , labels_row = names(new_genesets_ens1.C6), fontsize_row = 5, labels_col = numbers_patients, fontsize = 7) 
```

```{r}
#saveRDS(adj.enrichment_matrix.ens.C6, "/Users/paulaverdung/Desktop/Uni/Bachelor MoBi/Info/Bioinfoprojekt/NES.enrichment_matrix.C6.adj.ens.RDS")
```
We next try this for all C6 genesets (187):
```{r}
ensembl = useEnsembl(biomart = "genes", dataset = "hsapiens_gene_ensembl")
ensembl_lookup = getBM(attributes = c("ensembl_gene_id", "external_gene_name"),
                       mart = ensembl)
new_genesets_ens.C6.all = sapply(1:length(C6.all.genesets), function(x) {
  ensembl_lookup$ensembl_gene_id[which(ensembl_lookup$external_gene_name %in% C6.all.genesets[[x]])]
})
names(new_genesets_ens.C6.all) = names(C6.all.genesets)
```
GSEA:
```{r}
NES.list.patients.all.C6 = list()
for (patient in names(named.numericals.patients.ens)){
  stats.gsea.patients.ens = named.numericals.patients.ens[[patient]]
  gsea_patient_res.ens = fgseaSimple(
    stats = stats.gsea.patients.ens,
    nperm = 100,
    pathways = new_genesets_ens.C6.all
  )
  NES.list.patients.all.C6[[patient]] = gsea_patient_res.ens$NES
}
```
Enrichment matrix:
```{r}
enrichment_matrix.all.C6 = as.matrix(NES.list.patients.all.C6[[1]])
for (i in (2:58)){
  vec.ens.C6.all = as.matrix(NES.list.patients.all.C6[[i]])
  enrichment_matrix.all.C6 = cbind(enrichment_matrix.all.C6, vec.ens.C6.all)
}
rownames(enrichment_matrix.all.C6) = names(new_genesets_ens.C6.all)
colnames(enrichment_matrix.all.C6) = numbers_patients
```
Heatmap:
```{r}
heatm.C6.all = pheatmap(enrichment_matrix.all.C6, cluster_rows = , labels_row = names(new_genesets_ens.C6.all), fontsize_row = 5, labels_col = numbers_patients, fontsize = 7) 
```
```{r}
pval.list.patients.all.C6 = list()
for (patient in names(named.numericals.patients.ens)){
  stats.gsea.patients.ens.C6 = named.numericals.patients.ens[[patient]]
  gsea_patient_res.ens.C6 = fgseaSimple(
    stats = stats.gsea.patients.ens.C6,
    nperm = 100,
    pathways = new_genesets_ens.C6.all
  )
  pval.list.patients.all.C6[[patient]] = gsea_patient_res.ens.C6$pval
}
```
Compute and plot adjusted heatmap with p val = 0.15
```{r}
#store the adjusted p vals in a matrix of same format as the NES matrix
pval_matrix.all.C6 = as.matrix(pval.list.patients.all.C6[[1]])
for (i in (2:58)){
  p.vec.ens.C6 = as.matrix(pval.list.patients.all.C6[[i]])
  pval_matrix.all.C6 = cbind(pval_matrix.all.C6, p.vec.ens.C6)
}
#find positions in which p-values are above certain threshhold
pval.t.all.C6 = 0.15
nsig.NES.all.C6 = which(pval_matrix.all.C6 > pval.t.all.C6)
NES.enrichment_matrix.C6.adj.all = enrichment_matrix.all.C6
for (e in nsig.NES.all.C6){
  NES.enrichment_matrix.C6.adj.all[e] = 0
}
```
```{r}
adj.heatm.patietns.gsea.all.C6 = pheatmap(NES.enrichment_matrix.C6.adj.all, cluster_rows = , labels_row = names(C6.all.genesets), fontsize_row = 5, labels_col = numbers_patients, fontsize = 7) 
```
Select 30 most variant genesets
```{r}
row_variances <- apply(NES.enrichment_matrix.C6.adj.all, 1, var)
top_30_genesets <- order(row_variances, decreasing = TRUE)[1:30]
top_30_geneset_matrix <- NES.enrichment_matrix.C6.adj.all[top_30_genesets, ]
print(dim(top_30_geneset_matrix))
```
Plot cropped matrix 
```{r}
cropped.heatm.patietns.gsea.all.C6 = pheatmap(top_30_geneset_matrix, cluster_rows = , labels_row = names(top_30_genesets), fontsize_row = 5, labels_col = numbers_patients, fontsize = 7, color = colorRampPalette(c("dodgerblue", "white", "brown1"))(256), legend_labels = list(scale = "NES")) 
```
```{r}
#saveRDS(top_30_geneset_matrix, "/Users/paulaverdung/Desktop/Uni/Bachelor MoBi/Info/Bioinfoprojekt/top_30_geneset_matrix.RDS")
```
```{r}
library(pheatmap)
library(grid)

# Create the heatmap and draw it
cropped.heatm.patietns.gsea.all.C6 <- pheatmap(
  top_30_geneset_matrix, 
  cluster_rows = T,  # assuming you don't want to cluster rows
  labels_row = names(top_30_genesets), 
  fontsize_row = 5, 
  labels_col = numbers_patients, 
  fontsize = 7, 
  color = colorRampPalette(c("dodgerblue", "white", "brown1"))(256)
)

# Add the NES label above the legend
grid.text("NES", 
          x = unit(0.97, "npc"),  # Adjust the x position as needed
          y = unit(0.9, "npc"),   # Adjust the y position as needed
          gp = gpar(fontsize = 10))
```
```{r}
```
