```{r}
BiocManager::install("fgsea")
install.packages("msigdbr")
library(msigdbr)
library(tidyverse)
library(clusterProfiler)
library(RColorBrewer)
library(fgsea)
packageVersion("msigdbr")
```
First I created a data frame from the FoldChange values and then ordered it by V1 (fold change values) in descending order. Then I extracted the foldchange values into a list of named numerics (fold_change values matched to ensemble id):
```{r}
fold_change_df = df_FoldChange_ohne_NA
fold_change_df = fold_change_df[order(-fold_change_df$V1),]

#names_genes_fold_change = rownames(fold_change_df)
fold_change_ranked_list = fold_change_df$V1
names(fold_change_ranked_list) = rownames(fold_change_df)
```

```{r}
plot(fold_change_ranked_list)
```
The problem we encounter is that the gene sets only contain the gene name or entrez id, not the ensemble id, so we need to somehow rename the genes in the list. I would change them to the gene names as they are also represented in the gene sets by Hermann:
```{r}
library('biomaRt')
# Connect to BioMart
mart <- useMart("ensembl", dataset = "hsapiens_gene_ensembl")

# Get gene symbols for ensemble IDs
getGeneSymbols <- function(ensemble_ids, mart) {
  result <- getBM(attributes = c("ensembl_gene_id", "external_gene_name"),
                  filters = "ensembl_gene_id",
                  values = ensemble_ids,
                  mart = mart)
  return(result)
}

# Example list of ensemble IDs
ensemble_ids <- names(fold_change_ranked_list)

# Get gene symbols for ensemble IDs
gene_symbols <- getGeneSymbols(ensemble_ids, mart)
```
Now we need to link the gene symbols to the list (rename the named numericals):
```{r}
fold_change_ranked_list_gn = fold_change_ranked_list
names(fold_change_ranked_list_gn) = gene_symbols$external_gene_name
fold_change_ranked_list_gn
```
Run gene set enrichment analysis (GSEA via fgsea) across all gene sets from hermann (hallmark gene sets):
```{r}
gsea_LUAD_res = fgseaSimple(stats = fold_change_ranked_list_gn,
                      pathways = new_genesets,
                      nperm = 10000,
                      minSize = 1,
                      scoreType = "std",
                      maxSize = length(fold_change_ranked_list_gn)-1,
                      )
gsea_LUAD_res
````
```{r}
#saveRDS(gsea_LUAD_res, "/Users/paulaverdung/Desktop/Uni/Bachelor MoBi/Info/Bioinfoprojekt/gsea_LUAD_res.RDS")
```
Plotting the GSEA results in a barplot:
```{r}
# Plot GSEA results using barplot
barplot(gsea_LUAD_res$NES, 
        names.arg = gsea_LUAD_res$pathway, 
        las = 2,
        cex.names = 0.45,
        col = ifelse(gsea_LUAD_res$padj < 0.05, "red", "gray"),
        main = "GSEA results hallmark genesets",
        xlab = "Pathway",
        ylab = "Normalized Enrichment Score (NES)",
        )
```
Now I want to select the leading edge genes of the significantly upregulated and downregulated gene sets:
```{r}
leading_edge_hermann_a = gsea_LUAD_res$leadingEdge
#print(leading_edge1)
names(leading_edge_hermann_a) = gsea_LUAD_res$pathway
#select significantly upregulated genes 
leading_edge_hermann_up = leading_edge_hermann_a[which(gsea_LUAD_res$padj <= 0.05 & gsea_LUAD_res$NES > 0)]
#print(leading_edge_hermann_up)
#select significantly downregulated genes
leading_edge_hermann_down = leading_edge_hermann_a[which(gsea_LUAD_res$padj <= 0.05 & gsea_LUAD_res$NES < 0)]
#print(leading_edge_hermann_down)
```

Lookig for interesting pathways:
```{r}
#BiocManager::install("clusterProfiler", version = "3.18")
#BiocManager::install("pathview")
#BiocManager::install("enrichplot")

library(clusterProfiler)
library(enrichplot)
library(org.Hs.eg.db)
```


```{r}
big_vector_leading_edge_hermann_up= unlist(leading_edge_hermann_up, use.names = F)

organism = "org.Hs.eg.db"

# getting gene ID from gene names 
up_table_leading_hermann_names_big_vector = getBM(filters = "external_gene_name", 
                     attributes = c("ensembl_gene_id", "external_gene_name", "description", "gene_biotype"), 
                     values = big_vector_leading_edge_hermann_up, 
                     mart = ensembl)
# Pathway-Analysis
leading_edge_hermann_up_plot = enrichGO(gene = up_table_leading_hermann_names_big_vector$ensembl_gene_id,
               OrgDb = organism,
               keyType = "ENSEMBL",
               ont = "BP",
               pvalueCutoff = 0.05)

dotplot(leading_edge_hermann_up_plot)
```

```{r}
big_vector_leading_edge_hermann_down= unlist(leading_edge_hermann_down, use.names = F)

organism = "org.Hs.eg.db"

# getting gene ID from gene names 
down_table_leading_hermann_names_big_vector = getBM(filters = "external_gene_name", 
                     attributes = c("ensembl_gene_id", "external_gene_name", "description", "gene_biotype"), 
                     values = big_vector_leading_edge_hermann_down, 
                     mart = ensembl)
# Pathway-Analysis
leading_edge_hermann_down_plot = enrichGO(gene = down_table_leading_hermann_names_big_vector$ensembl_gene_id,
               OrgDb = organism,
               keyType = "ENSEMBL",
               ont = "BP",
               pvalueCutoff = 0.05)

dotplot(leading_edge_hermann_down_plot)
```


Experiment with msigdb data sets (hallmarks):
```{r}
library("dplyr")
library("msigdbr")
library("tidyverse")
H = msigdbr(species = "Homo sapiens", category = "H")
str(H)

#class(hallmark_gene_sets)
#we need to reformat the msigdbr output into a list of vectors (gene sets) with the corresponding genes in them

#very important code for GSEA with msigdb
H.genesymbols.ls = H %>%
  dplyr::select(gs_name, gene_symbol) %>%
  group_by(gs_name) %>%
  summarise(all.genes = list(unique(gene_symbol))) %>%
  deframe()
#GSEA on the msigdb gene sets
fgsea_res_msigdb = fgseaSimple(pathways = H.genesymbols.ls,
                         stats = fold_change_ranked_list_gn,
                         nperm = 10000,
                         minSize = 1,
                         maxSize = length(fold_change_ranked_list_gn) - 1)
```

Plotting the results:
```{r}
# Plot GSEA results using barplot
barplot(fgsea_res_msigdb$NES, 
        names.arg = fgsea_res_msigdb$pathway, 
        las = 2,
        cex.names = 0.3,
        col = ifelse(fgsea_res_msigdb$padj < 0.05, "red", "gray"),
        main = "GSEA Results",
        xlab = "Pathway",
        ylab = "Normalized Enrichment Score (NES)",
)
```
Now similar to above I will select the leading edge genes from all significantly down- or upregulated pathways:
```{r}
leading_edge_msigdb_a = gsea_LUAD_res$leadingEdge
#print(leading_edge_msigdb_a)
names(leading_edge_msigdb_a) = gsea_LUAD_res$pathway
#select significantly upregulated genes 
leading_edge_msigdb_up = leading_edge_msigdb_a[which(fgsea_res_msigdb$padj <= 0.05 & fgsea_res_msigdb$NES > 0)]
#print(leading_edge_hermann_up)
#select significantly downregulated genes
leading_edge_msigdb_down = leading_edge_msigdb_a[which(fgsea_res_msigdb$padj <= 0.05 & fgsea_res_msigdb$NES < 0)]
#print(leading_edge_msigdb_down)
```

Looking for interesting pathways: 
```{r}
big_vector_leading_edge_msigdb_up = unlist(leading_edge_msigdb_up, use.names = F)

organism = "org.Hs.eg.db"

# getting gene ID from gene names 
up_table_leading_msigdb_names_big_vector = getBM(filters = "external_gene_name", 
                     attributes = c("ensembl_gene_id", "external_gene_name", "description", "gene_biotype"), 
                     values = big_vector_leading_edge_msigdb_up, 
                     mart = ensembl)
# Pathway-Analysis
leading_edge_msigdb_up_plot = enrichGO(gene = up_table_leading_msigdb_names_big_vector$ensembl_gene_id,
               OrgDb = organism,
               keyType = "ENSEMBL",
               ont = "BP",
               pvalueCutoff = 0.05)

dotplot(leading_edge_msigdb_up_plot)
```

```{r}
big_vector_leading_edge_msigdb_down = unlist(leading_edge_msigdb_down, use.names = F)

organism = "org.Hs.eg.db"

# getting gene ID from gene names 
down_table_leading_msigdb_names_big_vector = getBM(filters = "external_gene_name", 
                     attributes = c("ensembl_gene_id", "external_gene_name", "description", "gene_biotype"), 
                     values = big_vector_leading_edge_msigdb_down, 
                     mart = ensembl)
# Pathway-Analysis
leading_edge_msigdb_down_plot = enrichGO(gene = down_table_leading_msigdb_names_big_vector$ensembl_gene_id,
               OrgDb = organism,
               keyType = "ENSEMBL",
               ont = "BP",
               pvalueCutoff = 0.05)

dotplot(leading_edge_msigdb_down_plot)
```


Import fold change data frame:
```{r}
#fold_change_df = t(do.call(cbind, FoldChange))
#fold_change_df = as.data.frame(fold_change_df)
#gene_ids_GSEA = rownames(fold_change_df)
#fold_change_df2 = cbind(gene_ids_GSEA, fold_change_df)
```
Ranking of fold change frame:
```{r}
#V1_sort = fold_change_df2[order(-fold_change_df2$V1), ]
```
Alternative (returns large numeric with ordered gene ids):
```{r}
#fold_change_df_alt = df_FoldChange_ohne_NA$V1
#names(fold_change_df_alt) = rownames(df_FoldChange_ohne_NA)

#V1_sort_alt = fold_change_df_alt[order(-fold_change_df)]
```