#GSEA (LUAD) 
## PCA: GSEA (adjusted values, including new gensets)

```{r}
# adjust dataframe 
Patienten_ID_LUAD <- colnames(luad.tumor_filtered_proteins)
colnames(NES.enrichment_matrix.C6.adj.ens) <- Patienten_ID_LUAD
```

```{r}
# load needed libraries
library(irlba)
library(ggplot2)

# transposing matrix
NES.enrichment_matrix_t <- t(NES.enrichment_matrix.C6.adj.ens)

# calculate PCA
pca_NES.enrichment_matrix <- irlba(NES.enrichment_matrix_t, nv = 10)
```

```{r}
# check PCA results
pca_NES.enrichment_matrix
```

```{r}
# adjust the structure to a prcomp-like structure
pca_prcomp_NES.enrichment_matrix <- list()

pca_prcomp_NES.enrichment_matrix$sdev <- sqrt(pca_NES.enrichment_matrix$d^2 / (nrow(NES.enrichment_matrix_t) - 1))

pca_prcomp_NES.enrichment_matrix$rotation <- pca_NES.enrichment_matrix$v

pca_prcomp_NES.enrichment_matrix$x <- NES.enrichment_matrix_t %*% pca_NES.enrichment_matrix$v

pca_prcomp_NES.enrichment_matrix$center <- colMeans(NES.enrichment_matrix_t)

pca_prcomp_NES.enrichment_matrix$scale <- apply(NES.enrichment_matrix_t, 2, sd)

class(pca_prcomp_NES.enrichment_matrix) <- "prcomp"
```

```{r}
# combine PCA results of first two dimensions in a new data frame 
plot_GSEA <- data.frame(PC1 = pca_prcomp_NES.enrichment_matrix$x[, 1], PC2 = pca_prcomp_NES.enrichment_matrix$x[, 2])

ggplot(plot_GSEA, aes(x = PC1, y = PC2)) +
  geom_point(size = 1) +
  labs(title = "PCA Scatterplot",
       x = "PC1",
       y = "PC2") +
  theme_minimal() +
  theme(legend.title = element_blank())
```

```{r}
# combine PCA results & clinical annotations in a new data frame 
plot_GSEA_tumortypes <- data.frame(PC1 = pca_prcomp_NES.enrichment_matrix$x[, 1], PC2 = pca_prcomp_NES.enrichment_matrix$x[, 2], Tumorstage = luad.annot$ajcc_pathologic_tumor_stage)

# plot data
ggplot(plot_GSEA_tumortypes, aes(x = PC1, y = PC2, color = Tumorstage)) +
  geom_point(size = 1) +
  labs(title = "PCA Scatterplot nach Tumorstage",
       x = "PC1",
       y = "PC2") +
  theme_minimal() +
  theme(legend.title = element_blank())
```

## UMAP: GSEA (adjusted values, including new genesets)

```{r}
library(umap)

# compute a umap
umap_NES.enrichment_matrix <- umap(NES.enrichment_matrix.C6.adj.ens)

# check dimension 
print(dim(umap_NES.enrichment_matrix$layout))  # should be n x 2  
```

```{r}
# combine data for plotting 
umap_gsea <- data.frame(
  UMAP1 = umap_NES.enrichment_matrix$layout[, 1],
  UMAP2 = umap_NES.enrichment_matrix$layout[, 2])



# plot umap results
ggplot(umap_gsea, aes(x = UMAP1, y = UMAP2)) +
  geom_point(size = 1) +
  labs(title = "UMAP Plot",
       x = "UMAP Dimension 1",
       y = "UMAP Dimension 2") +
  theme_minimal() 
```

## all genesets (PCA followed by UMAP)

```{r}
# combine data
NES.enrichment_matrix_comb <- rbind(NES.enrichment_matrix.C6.ens, NES.enrichment_matrix.ens)

# adjust dataframe
Patienten_ID_LUAD <- colnames(luad.tumor_filtered_proteins)
colnames(NES.enrichment_matrix_comb) <- Patienten_ID_LUAD
```

```{r}
# load libraries
library(irlba)
library(ggplot2)

# transpose matrix
NES.enrichment_matrix_comb_t <- t(NES.enrichment_matrix_comb)

# compute PCA
pca_NES.enrichment_matrix_comb <- irlba(NES.enrichment_matrix_comb_t, nv = 10)
```

```{r}
# adjust structure to a prcomp-like one
pca_prcomp_NES.enrichment_matrix_comb <- list()

pca_prcomp_NES.enrichment_matrix_comb$sdev <- sqrt(pca_NES.enrichment_matrix_comb$d^2 / (nrow(NES.enrichment_matrix_comb_t) - 1))

pca_prcomp_NES.enrichment_matrix_comb$rotation <- pca_NES.enrichment_matrix_comb$v

pca_prcomp_NES.enrichment_matrix_comb$x <- NES.enrichment_matrix_comb_t %*% pca_NES.enrichment_matrix_comb$v

pca_prcomp_NES.enrichment_matrix_comb$center <- colMeans(NES.enrichment_matrix_comb_t)

pca_prcomp_NES.enrichment_matrix_comb$scale <- apply(NES.enrichment_matrix_comb_t, 2, sd)

class(pca_prcomp_NES.enrichment_matrix_comb) <- "prcomp"
```

```{r}
# extract scores (PCA-coordinates of data points) 
pca_NES.enrichment_matrix_comb_scores <- pca_NES.enrichment_matrix_comb$u

# check dimension of pca results
print(dim(pca_NES.enrichment_matrix_comb_scores))  
```

```{r}
# compute umap on pca results
umap_pca_NES.enrichment_matrix_comb <- umap(pca_NES.enrichment_matrix_comb_scores)

# check dimension 
print(dim(umap_pca_NES.enrichment_matrix_comb$layout))  # should be n x 2
```

```{r}
# combine needed data
umap_pca_comb <- data.frame(
  UMAP1 = umap_pca_NES.enrichment_matrix_comb$layout[, 1],
  UMAP2 = umap_pca_NES.enrichment_matrix_comb$layout[, 2])

# plot umap results
ggplot(umap_pca_comb, aes(x = UMAP1, y = UMAP2)) +
  geom_point(size = 1) +
  labs(title = "UMAP plot ",
       x = "UMAP dimension 1",
       y = "UMAP dimension 2") +
  theme_minimal() +
  theme(legend.position = "right")  
```

```{r}
# combine needed data
umap_pca_comb <- data.frame(
  UMAP1 = umap_pca_NES.enrichment_matrix_comb$layout[, 1],
  UMAP2 = umap_pca_NES.enrichment_matrix_comb$layout[, 2],
  Tumorstage = luad.annot$ajcc_pathologic_tumor_stage
)

# plot umap results
ggplot(umap_pca_comb, aes(x = UMAP1, y = UMAP2, color = Tumorstage)) +
  geom_point(size = 1) +
  labs(title = "UMAP Plot nach Tumorstage",
       x = "UMAP Dimension 1",
       y = "UMAP Dimension 2") +
  theme_minimal() +
  theme(legend.position = "right")  
```

# top 30 genesets

## PCA followed by UMAP

```{r}
# load libraries
library(irlba)
library(ggplot2)

# transpose matrix
top_30_geneset_matrix_t <- t(top_30_geneset_matrix)

# compute PCA
pca_top_30_geneset_matrix <- irlba(top_30_geneset_matrix_t, nv = 10)
```

```{r}
# adjust structure to a prcomp-like structure
pca_prcomp_top_30_geneset_matrix <- list()

pca_prcomp_top_30_geneset_matrix$sdev <- sqrt(pca_top_30_geneset_matrix$d^2 / (nrow(top_30_geneset_matrix_t) - 1))

pca_prcomp_top_30_geneset_matrix$rotation <- pca_top_30_geneset_matrix$v

pca_prcomp_top_30_geneset_matrix$x <- top_30_geneset_matrix_t %*% pca_top_30_geneset_matrix$v

pca_prcomp_top_30_geneset_matrix$center <- colMeans(top_30_geneset_matrix_t)

pca_prcomp_top_30_geneset_matrix$scale <- apply(top_30_geneset_matrix_t, 2, sd)

class(pca_prcomp_top_30_geneset_matrix) <- "prcomp"
```

```{r}
# extract scores 
pca_top_30_geneset_matrix_scores <- pca_top_30_geneset_matrix$u

# check dimension 
print(dim(pca_top_30_geneset_matrix_scores))  
```

```{r}
# compute umap on pca results
umap_pca_top_30_geneset_matrix <- umap(pca_top_30_geneset_matrix_scores)

# check dimension
print(dim(umap_pca_top_30_geneset_matrix$layout))  # should be n x 2 
```

```{r}
# combine needed data
umap_pca_top30 <- data.frame(
  UMAP1 = umap_pca_top_30_geneset_matrix$layout[, 1],
  UMAP2 = umap_pca_top_30_geneset_matrix$layout[, 2])

# plot umap results
ggplot(umap_pca_top30, aes(x = UMAP1, y = UMAP2)) +
  geom_point(size = 1) +
  labs(title = "UMAP plot: top 30 gene sets",
       x = "UMAP dimension 1",
       y = "UMAP dimension 2") +
  theme_minimal() +
  theme(legend.position = "right")  
```

## UMAP

```{r}
library(umap)

# compute a umap
umap_top_30_geneset_matrix <- umap(top_30_geneset_matrix)

# check dimension of results
print(dim(umap_top_30_geneset_matrix$layout))  # should be n x 2
```

```{r}
# combine data 
umap_top30 <- data.frame(
  UMAP1 = umap_top_30_geneset_matrix$layout[, 1],
  UMAP2 = umap_top_30_geneset_matrix$layout[, 2])

# plot results
ggplot(umap_top30, aes(x = UMAP1, y = UMAP2)) +
  geom_point(size = 1) +
  labs(title = "UMAP Plot",
       x = "UMAP Dimension 1",
       y = "UMAP Dimension 2") +
  theme_minimal() 
```


# FeaturePlot: most variant genesets of hallmark genesets

closer look at: Angio_ALL, Angio_P_Reg, Immune_AACR, Metas_NEG, Prol_ERK, Metas_AACR, Metas_ALL, Genome_AACR, Inflam_AACR

```{r}
# extract enrichment scores of special genesets / pathways
scores <- NES.enrichment_matrix.ens[1:2, ]
scores <- t(scores)
patient <-rownames(umap_luad.tumor_filtered_proteins_t$layout)
rownames(scores) <- patient
scores

# combining umap results and enrichment scores
umap_12 <- umap_luad.tumor_filtered_proteins_t$layout
umap <- cbind(umap_12, scores)
umap <- as.data.frame(umap)


# create UMAP-Plot (colored tumorstages)
ggplot(umap, aes(x = V1, y = V2, color = Angio_ALL)) +
  geom_point(size = 2) +
  labs(title = "UMAP Plot with colored tumorstages",
       x = "UMAP Dimension 1",
       y = "UMAP Dimension 2") +
  scale_color_gradient(low = "dodgerblue1", high = "brown1", name = "Angio_ALL") +
  theme_minimal() +
  theme(legend.position = "right")

```

```{r}
# extract enrichment scores of special genesets / pathways
scores <- NES.enrichment_matrix.ens[5:6, ]
scores <- t(scores)
patient <-rownames(umap_luad.tumor_filtered_proteins_t$layout)
rownames(scores) <- patient
scores

# combining umap results and enrichment scores
umap_12 <- umap_luad.tumor_filtered_proteins_t$layout
umap <- cbind(umap_12, scores)
umap <- as.data.frame(umap)


# create UMAP-Plot (colored tumorstages)
ggplot(umap, aes(x = V1, y = V2, color = Angio_P_Reg)) +
  geom_point(size = 2) +
  labs(title = "UMAP Plot with colored tumorstages",
       x = "UMAP Dimension 1",
       y = "UMAP Dimension 2") +
  scale_color_gradient(low = "dodgerblue1", high = "brown1", name = "Angio_P_Reg") +
  theme_minimal() +
  theme(legend.position = "right")
```

```{r}
# extract enrichment scores of special genesets / pathways
scores <- NES.enrichment_matrix.ens[15:16, ]
scores <- t(scores)
patient <-rownames(umap_luad.tumor_filtered_proteins_t$layout)
rownames(scores) <- patient
scores

# combining umap results and enrichment scores
umap_12 <- umap_luad.tumor_filtered_proteins_t$layout
umap <- cbind(umap_12, scores)
umap <- as.data.frame(umap)


# create UMAP-Plot (colored tumorstages)
ggplot(umap, aes(x = V1, y = V2, color = Immune_AACR)) +
  geom_point(size = 2) +
  labs(title = "UMAP Plot with colored tumorstages",
       x = "UMAP Dimension 1",
       y = "UMAP Dimension 2") +
  scale_color_gradient(low = "dodgerblue1", high = "brown1", name = "Immune_AACR") +
  theme_minimal() +
  theme(legend.position = "right")
```

```{r}
# extract enrichment scores of special genesets / pathways
scores <- NES.enrichment_matrix.ens[32:33, ]
scores <- t(scores)
patient <-rownames(umap_luad.tumor_filtered_proteins_t$layout)
rownames(scores) <- patient
scores

# combining umap results and enrichment scores
umap_12 <- umap_luad.tumor_filtered_proteins_t$layout
umap <- cbind(umap_12, scores)
umap <- as.data.frame(umap)


# create UMAP-Plot (colored tumorstages)
ggplot(umap, aes(x = V1, y = V2, color = Metas_NEG)) +
  geom_point(size = 2) +
  labs(title = "UMAP Plot with colored tumorstages",
       x = "UMAP Dimension 1",
       y = "UMAP Dimension 2") +
  scale_color_gradient(low = "dodgerblue1", high = "brown1", name = "Metas_NEG") +
  theme_minimal() +
  theme(legend.position = "right")
```

```{r}
# extract enrichment scores of special genesets / pathways
scores <- NES.enrichment_matrix.ens[40:41, ]
scores <- t(scores)
patient <-rownames(umap_luad.tumor_filtered_proteins_t$layout)
rownames(scores) <- patient
scores

# combining umap results and enrichment scores
umap_12 <- umap_luad.tumor_filtered_proteins_t$layout
umap <- cbind(umap_12, scores)
umap <- as.data.frame(umap)


# create UMAP-Plot (colored tumorstages)
ggplot(umap, aes(x = V1, y = V2, color = Prol_ERK)) +
  geom_point(size = 2) +
  labs(title = "UMAP Plot with colored tumorstages",
       x = "UMAP Dimension 1",
       y = "UMAP Dimension 2") +
  scale_color_gradient(low = "dodgerblue1", high = "brown1", name = "Prol_ERK") +
  theme_minimal() +
  theme(legend.position = "right")
```

```{r}
# extract enrichment scores of special genesets / pathways
scores <- NES.enrichment_matrix.ens[30:31, ]
scores <- t(scores)
patient <-rownames(umap_luad.tumor_filtered_proteins_t$layout)
rownames(scores) <- patient
scores

# combining umap results and enrichment scores
umap_12 <- umap_luad.tumor_filtered_proteins_t$layout
umap <- cbind(umap_12, scores)
umap <- as.data.frame(umap)


# create UMAP-Plot (colored tumorstages)
ggplot(umap, aes(x = V1, y = V2, color = Metas_AACR)) +
  geom_point(size = 2) +
  labs(title = "UMAP Plot with colored tumorstages",
       x = "UMAP Dimension 1",
       y = "UMAP Dimension 2") +
  scale_color_gradient(low = "dodgerblue1", high = "brown1", name = "Metas_AACR") +
  theme_minimal() +
  theme(legend.position = "right")
```

```{r}
# extract enrichment scores of special genesets / pathways
scores <- NES.enrichment_matrix.ens[30:31, ]
scores <- t(scores)
patient <-rownames(umap_luad.tumor_filtered_proteins_t$layout)
rownames(scores) <- patient
scores

# combining umap results and enrichment scores
umap_12 <- umap_luad.tumor_filtered_proteins_t$layout
umap <- cbind(umap_12, scores)
umap <- as.data.frame(umap)


# create UMAP-Plot (colored tumorstages)
ggplot(umap, aes(x = V1, y = V2, color = Metas_ALL)) +
  geom_point(size = 2) +
  labs(title = "UMAP Plot with colored tumorstages",
       x = "UMAP Dimension 1",
       y = "UMAP Dimension 2") +
  scale_color_gradient(low = "dodgerblue1", high = "brown1", name = "Metas_ALL") +
  theme_minimal() +
  theme(legend.position = "right")
```

```{r}
# extract enrichment scores of special genesets / pathways
scores <- NES.enrichment_matrix.ens[9:10, ]
scores <- t(scores)
patient <-rownames(umap_luad.tumor_filtered_proteins_t$layout)
rownames(scores) <- patient
scores

# combining umap results and enrichment scores
umap_12 <- umap_luad.tumor_filtered_proteins_t$layout
umap <- cbind(umap_12, scores)
umap <- as.data.frame(umap)


# create UMAP-Plot (colored tumorstages)
ggplot(umap, aes(x = V1, y = V2, color = Genome_AACR)) +
  geom_point(size = 2) +
  labs(title = "UMAP Plot with colored tumorstages",
       x = "UMAP Dimension 1",
       y = "UMAP Dimension 2") +
  scale_color_gradient(low = "dodgerblue1", high = "brown1", name = "Genome_AACR") +
  theme_minimal() +
  theme(legend.position = "right")
```

```{r}
# extract enrichment scores of special genesets / pathways
scores <- NES.enrichment_matrix.ens[20:21, ]
scores <- t(scores)
patient <-rownames(umap_luad.tumor_filtered_proteins_t$layout)
rownames(scores) <- patient
scores

# combining umap results and enrichment scores
umap_12 <- umap_luad.tumor_filtered_proteins_t$layout
umap <- cbind(umap_12, scores)
umap <- as.data.frame(umap)


# create UMAP-Plot (colored tumorstages)
ggplot(umap, aes(x = V1, y = V2, color = Inflam_AACR)) +
  geom_point(size = 2) +
  labs(title = "UMAP Plot with colored tumorstages",
       x = "UMAP Dimension 1",
       y = "UMAP Dimension 2") +
  scale_color_gradient(low = "dodgerblue1", high = "brown1", name = "Inflam_AACR") +
  theme_minimal() +
  theme(legend.position = "right")
```

# all genesets (hallmarks & C6 combined)

```{r}
NES.enrichment_matrix_comb <- rbind(NES.enrichment_matrix.C6.ens, NES.enrichment_matrix.ens)
```

```{r}
# extract enrichment scores of special genesets / pathways
scores <- NES.enrichment_matrix_comb[10:11, ]
scores <- t(scores)
patient <-rownames(umap_luad.tumor_filtered_proteins_t$layout)
rownames(scores) <- patient
scores

# combining umap results and enrichment scores
umap_12 <- umap_luad.tumor_filtered_proteins_t$layout
umap <- cbind(umap_12, scores)
umap <- as.data.frame(umap)


# create UMAP-Plot (colored tumorstages)
ggplot(umap, aes(x = V1, y = V2, color = MTOR_UP.N4.V1_UP)) +
  geom_point(size = 2) +
  labs(title = "UMAP Plot with colored tumorstages",
       x = "UMAP Dimension 1",
       y = "UMAP Dimension 2") +
  scale_color_gradient(low = "dodgerblue1", high = "brown1", name = "MTOR_UP.N4.V1_UP
") +
  theme_minimal() +
  theme(legend.position = "right")
```
