Checking for missing values (NA):

```{r}
total_na_count_exp = sum(is.na(tcga_exp))
print(total_na_count_exp)

total_na_count_annot = sum(is.na(tcga_annot))
print(total_na_count_annot)

total_na_count_tumor_norm = sum(is.na(tcga_tumor_norm))
print(total_na_count_tumor_norm)

total_na_count_genesets = sum(is.na(genesets))
print(total_na_count_genesets)
```

Only the tcga_annot dataset contains missing values.

Installation BiomaRt:

```{r}
if (!require("BiocManager", quietly = TRUE))
    install.packages("BiocManager")

BiocManager::install("biomaRt")
```


Extraction of protein-coding genes: 

```{r}
# install.packages(c("biomaRt", "AnnotationDbi")) if necessary 
library(biomaRt)
library(AnnotationDbi)

mart <- useMart("ensembl")
ensembl = useMart("ensembl", dataset = "hsapiens_gene_ensembl")

# list of all humane genes and their attributes 
genes = getBM(attributes = c("ensembl_gene_id", "gene_biotype"), mart = ensembl)

# only protein-coding genes
protein_coding_genes = subset(genes, gene_biotype == "protein_coding")

# attention --> this code inserts only an empty column at the end oft the dara record 
tcga_exp$ensembl_gene_id = sub("\\..*", "", rownames(tcga_exp))

# Filtering
tcga_exp_filtered_proteins <- tcga_exp[tcga_exp$ensembl_gene_id %in% protein_coding_genes$ensembl_gene_id, ]

# rownames are overwritten with gene IDs 
rownames(tcga_exp_filtered_proteins) <- tcga_exp_filtered_proteins[, ncol(tcga_exp_filtered_proteins)]

# remove last column 
tcga_exp_filtered_proteins <- tcga_exp_filtered_proteins[, -ncol(tcga_exp_filtered_proteins)]

dim(tcga_exp)
dim(tcga_exp_filtered_proteins)
#print(nrow(protein_coding_genes))
```

```{r}
tcga_exp_begrenzt1 <- tcga_exp_filtered_proteins[1:10000, ]
dim(tcga_exp_begrenzt1)

tcga_exp_begrenzt_2 <- tcga_exp_filtered_proteins[10001:19579, ]
dim(tcga_exp_begrenzt_2)
```

protein-coding genes are plotted in a mean-variance plot and sorted out according to a threshold: for the calculations, the data set must be divided into two parts (otherwise the computer vannot provide the computing power)

```{r}
# Mittelwert jeder Zeile berechnen - Teil 1
mean_Zeilen_tcga_exp_begrenzt1 <- apply(tcga_exp_begrenzt1, 1, mean)

# Varianz jeder Zeile berechnen
varianzen_Zeilen_tcga_exp_begrenzt1 <- apply(tcga_exp_begrenzt1, 1, var)


# Mittelwert jeder Zeile berechnen - Teil 2
mean_Zeilen_tcga_exp_begrenzt_2 <- apply(tcga_exp_begrenzt_2, 1, mean)

# Varianz jeder Zeile berechnen
varianzen_Zeilen_tcga_exp_begrenzt_2 <- apply(tcga_exp_begrenzt_2, 1, var)

```

after the calculation, the parts are now reassembled:

```{r}
alle_means_tcga_exp_filtered_proteins = c(mean_Zeilen_tcga_exp_begrenzt1, mean_Zeilen_tcga_exp_begrenzt_2)
alle_varianzen_tcga_exp_filtered_proteins = c(varianzen_Zeilen_tcga_exp_begrenzt1, varianzen_Zeilen_tcga_exp_begrenzt_2)
```

The previously calculated values are now to be plotted ub a mean-variance diagram. Two thresholds are calculated for this, one for the top 5% and the other for exactly 50% of the data. The data points categorised by the threshilds are color-coded accordingly. 

```{r}
# calculating thresholds 
threshold_oben = quantile(alle_varianzen_tcga_exp_filtered_proteins, probs = 0.95)
top_5_prozent_varianzen = which(alle_varianzen_tcga_exp_filtered_proteins >= threshold_oben)

threshold_50 = quantile(alle_varianzen_tcga_exp_filtered_proteins, probs = 0.50)
bad_50_prozent_varianzen = which(alle_varianzen_tcga_exp_filtered_proteins <= threshold_50)

# mean-variance-plot
plot(alle_means_tcga_exp_filtered_proteins, alle_varianzen_tcga_exp_filtered_proteins, main = "Mean-Variance Plot", xlab = "Mean", ylab = "Variance", cex.lab = 1.5, pch = 19, cex = 0.1)

# color-coding of data points 
points(alle_means_tcga_exp_filtered_proteins[top_5_prozent_varianzen], alle_varianzen_tcga_exp_filtered_proteins[top_5_prozent_varianzen], col = "firebrick2", pch = 19, cex = 0.1)

points(alle_means_tcga_exp_filtered_proteins[bad_50_prozent_varianzen], alle_varianzen_tcga_exp_filtered_proteins[bad_50_prozent_varianzen], col = "dodgerblue2", pch = 19, cex = 0.1)

```


Final Data Set (proteins/variance filtered): top 50% of the protein-coding genes with the highest variance 
```{r}
top_50_prozent_varianzen = which(alle_varianzen_tcga_exp_filtered_proteins >= threshold_50)

#new working data set:
tcga_exp_proteins_shortened = tcga_exp_filtered_proteins[top_50_prozent_varianzen, ]
dim(tcga_exp_proteins_shortened)
```

The results of the filtering just carried out are illustrated again graphically below.

```{r}
varianzen_tcga_shortened = apply(tcga_exp_proteins_shortened, 1, var)
```

```{r}
install.packages("ggplot2")
```

Violin plot of the original data and the filtered / reduced data:

```{r}
library(ggplot2)

C1 = data.frame(value = alle_varianzen_tcga_exp_filtered_proteins, variable = " protein-coding genes")
C2 = data.frame(value = varianzen_tcga_shortened, variable = "reduced protein-coding genes")

df = rbind(C1, C2)
ggplot(df, aes(x = variable, y = value)) +
  geom_violin(scale = "width", adjust = 1, width= 0.5) +
  labs(y = "Variance", title = "Violinplot", x = NULL) + 
  geom_boxplot(width=0.03) + 
  geom_hline(yintercept = median(alle_varianzen_tcga_exp_filtered_proteins), color = "red") +
  geom_text(aes(0.5, median(alle_varianzen_tcga_exp_filtered_proteins), label = "q50", vjust = -1), col = "red") +
  theme(axis.text.x = element_text(size = 18),
   axis.title.y = element_text(size = 18))

```

A PCA is now applied to the filtered data set for the purposes of dimension reduction and visualisation.

```{r}
# installing irlba 
install.packages("irlba")

```

```{r}
# loading package 
library(irlba)

# calculating pca with the help of the function "irlba"
pca_resultat <- irlba(as.matrix(tcga_exp_proteins_shortened), nv = 10)
pca_resultat
```

```{r}
# proportion of standard deviation explained by each principal component is contained in the pca$d vector:
pca_resultat$d
```

```{r}
# plot the proportion of total variance explained by each PC
variance = (pca_resultat$d)^2
prop.variance = variance/sum(variance)
names(prop.variance) = 1:length(prop.variance)
barplot(prop.variance[1:20],ylab='Proportion of variance') # we only plot the first 10 PCs
```

```{r}
# computing the scores of the PCs
scores <- pca_resultat$v[, 1:2]

# create a scatter plot of the scores
plot(scores[, 1], scores[, 2], xlab = "PC1", ylab = "PC2", main = "PCA Scatterplot")

```


```{r}
# loading needed libraries
library(irlba)
library(ggplot2)

# transpose matrix so that the columns represent the patients 
tcga_exp_proteins_shortened_t <- t(tcga_exp_proteins_shortened)

# calculating a PCA for the transposed matrix
pca_resultat <- irlba(tcga_exp_proteins_shortened_t, nv = 10)

# for the following graphical representations, the PCA results need to be converted into a prcomp-like structure 
pca_prcomp <- list()
pca_prcomp$sdev <- sqrt(pca_resultat$d^2 / (nrow(tcga_exp_proteins_shortened_t) - 1))
pca_prcomp$rotation <- pca_resultat$v
pca_prcomp$x <- tcga_exp_proteins_shortened_t %*% pca_resultat$v
pca_prcomp$center <- colMeans(tcga_exp_proteins_shortened_t)
pca_prcomp$scale <- apply(tcga_exp_proteins_shortened_t, 2, sd)
class(pca_prcomp) <- "prcomp"

# creating a dataframe that combines the PCA results and the clinical annotations 
plot_data_tumortypes_t <- data.frame(PC1 = pca_prcomp$x[, 1], PC2 = pca_prcomp$x[, 2], Tumortyp = tcga_annot$cancer_type_abbreviation)

# create a scatter plot of the scores, where the data points are colored according to their tumortype
ggplot(plot_data_tumortypes_t, aes(x = PC1, y = PC2, color = Tumortyp)) +
  geom_point(size = 2) +
  labs(title = "PCA Scatterplot nach Tumortypen",
       x = "PC1",
       y = "PC2") +
  theme_minimal() +
  theme(legend.title = element_blank())

```

Visualisation of all individual loading of the two main principle components of the PCA:

```{r}
# loading needed libraries
library(irlba)
library(ggplot2)
library(reshape2)

# extract the loadings of the first two PCs 
loadings <- pca_resultat$u[, 1:2]
colnames(loadings) <- c("PC1", "PC2")

# create a DataFrame of the loadings 
loadings_df <- as.data.frame(loadings)
loadings_df$Variable <- rownames(loadings_df)

# reshaping data for ggplot2 function 
loadings_melt <- melt(loadings_df, id.vars = "Variable")

# create Loadings plot 
ggplot(loadings_melt, aes(x = Variable, y = value, fill = variable)) +
  geom_bar(stat = "identity", position = "dodge") +
  labs(title = "Loadings Plot of PCs",
       x = "Variablen",
       y = "Loadings",
       fill = "principle components") +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust=1))

```

```{r}
# extract loadings of two main principle components 
loadings <- pca_resultat$u[, 1:2]
colnames(loadings) <- c("PC1", "PC2")

# calculate and sort sum of the absolute vaules of the loadings 
loading_sums <- rowSums(abs(loadings))
top_20_indices <- order(loading_sums, decreasing = TRUE)[1:20]

# create dataframe of top 20 variables of the loadings d
loadings_df <- as.data.frame(loadings[top_20_indices, ])
loadings_df$Variable <- rownames(loadings_df)

# reshape data
loadings_melt <- melt(loadings_df, id.vars = "Variable")

# create Loadings Plot 
ggplot(loadings_melt, aes(x = Variable, y = value, fill = variable)) +
  geom_bar(stat = "identity", position = "dodge") +
  labs(title = "Loadings Plot der Hauptkomponenten (Top 20 Variablen)",
       x = "Variablen",
       y = "Loadings",
       fill = "Hauptkomponente") +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust=1))
```


UMAP defines number of expected nearest neighbours & minimum distance (how tightly UMAP packs points which are close together)

```{r}
install.packages("umap")
```

```{r}
# loading needed libraries
library(irlba)
library(umap)
library(ggplot2)

# transpose matrix so that rows represent the patients 
tcga_exp_proteins_shortened_t <- t(tcga_exp_proteins_shortened)

# calculating PCA
pca_result_umap <- irlba(tcga_exp_proteins_shortened_t, nv = 20)
```

```{r}
# checking results of the PCA 
pca_result_umap
```

```{r}
# extract the scores (PCA-coordinates of the data points)
pca_scores_umap <- pca_result_umap$u

# checking dimension of PCA-results
print(dim(pca_scores_umap))  # should be n x 20  (n = number of data points)
```

```{r}
# apply UMAP function to the resukts of the PCA 
umap_result <- umap(pca_scores_umap)

# check dimension of UMAP-results
print(dim(umap_result$layout))  # should be n x 2  (n = number of data points)
```

```{r}
plot(umap_result$layout[,1], umap_result$layout[,2],
     xlab = "UMAP Dimension 1", ylab = "UMAP Dimension 2",
     main = "UMAP Plot")
```


```{r}
# combine umap results and clinical annotations
umap_data <- data.frame(
  UMAP1 = umap_result$layout[, 1],
  UMAP2 = umap_result$layout[, 2],
  Tumortype = tcga_annot$cancer_type_abbreviation
)

# create UMAP-Plot
ggplot(umap_data, aes(x = UMAP1, y = UMAP2, color = Tumortype)) +
  geom_point(size = 2) +
  labs(title = "UMAP Plot according to tumortyps",
       x = "UMAP Dimension 1",
       y = "UMAP Dimension 2") +
  theme_minimal() +
  theme(legend.position = "right") 
```
