In this part of the project we want to perform a GSEA on each patient, comparing healthy (normal) tissue to unhealthy (LUAD) tissue. To do so we need to first calculate the fold change values of all genes across all LUAD patients.
```{r}
fold_changes_patients = luad.tumor_filtered_proteins / luad.norm_filtered_proteins
```
Now I want to convert the ensemble ids to gene names using bio mart filtering 
```{r}
library('biomaRt')
# Connect to BioMart
mart <- useMart("ensembl", dataset = "hsapiens_gene_ensembl")

# Get gene symbols for ensemble IDs
getGeneSymbols <- function(ensemble_ids_patients, mart) {
  result <- getBM(attributes = c("ensembl_gene_id", "external_gene_name"),
                  filters = "ensembl_gene_id",
                  values = ensemble_ids_patients,
                  mart = mart)
  return(result)
}

# Example list of ensemble IDs
ensemble_ids_patients <- rownames(fold_changes_patients)

# Get gene symbols for ensemble IDs
gene_symbols_patients <- getGeneSymbols(ensemble_ids_patients, mart)

#certain gene names doubled, so I had to remove them in order to create a data frame with the corresponding gene names
remove_rows = which((duplicated(gene_symbols_patients$external_gene_name)))
gs_cut = gene_symbols_patients[-remove_rows,]

fold_changes_patients_gn_cut = fold_changes_patients[-remove_rows,]
rownames(fold_changes_patients_gn_cut) = gs_cut$external_gene_name
fold_changes_patients_gn_cut
```
In order to more easily iterate over all patients I simply convert patients names to numbers counting from one onwards
```{r}
numbers_patients = c(1:length(colnames(fold_changes_patients_gn_cut)))

colnames(fold_changes_patients_gn_cut) = numbers_patients
```

Next we need to rank the fold change values for each patient by decreasing order and want to store the corresponding gene names in the same order for each patient in a named list:
```{r}
rank_list_patients = list()
patient = 1

for (patient in colnames(fold_changes_patients_gn_cut)){
  patient_fc = fold_changes_patients_gn_cut[, patient]
  names(patient_fc) = rownames(fold_changes_patients_gn_cut)
  rank_list_patients[[patient]] = (sort(patient_fc, decreasing = T))
}

```

Now we can compute a list in which we store the same lists but instead as named numericals
```{r}
named.numericals.patients = list()
for(patient in names(rank_list_patients)){
 named.num = unlist(rank_list_patients[patient])
 names(named.num) = names(rank_list_patients[[patient]])
 named.numericals.patients[[patient]] = named.num
}
```
Next we want to compute the GSEA on each patient
```{r}
NES.list.patients = list()
for (patient in names(named.numericals.patients)){
  stats.gsea.patients = named.numericals.patients[[patient]]
  gsea_patient_res = fgseaSimple(
    stats = stats.gsea.patients,
    nperm = 100,
    pathways = new_genesets
  )
  NES.list.patients[[patient]] = gsea_patient_res$NES
}
```
I also want to store the names of the gene_sets I calculated the NES for in the corresponding lists
```{r}
#for (i in names(NES.list.patients)){
 # names_new_genesets = names(new_genesets)
  #names(NES.list.patients[[i]]) = names_new_genesets
}
```
Last but not least I want to plot the results of the gsea using a heatmap
```{r}
enrichment_matrix = as.matrix(NES.list.patients[[1]])
for (i in (2:58)){
  vec = as.matrix(NES.list.patients[[i]])
  enrichment_matrix = cbind(enrichment_matrix, vec)
}
rownames(enrichment_matrix) = names(new_genesets)
colnames(enrichment_matrix) = numbers_patients
````
```{r}
heatm.patietns.gsea = pheatmap(enrichment_matrix, cluster_rows = , labels_row = names_new_genesets, fontsize_row = 5, labels_col = numbers_patients, fontsize = 7, color = colorRampPalette(c("dodgerblue", "white", "brown1"))(256))  #,color = colorRampPalette(brewer.pal(10, "RdBu"))(256)) 

```
```{r}
#saveRDS(enrichment_matrix, "/Users/paulaverdung/Desktop/Uni/Bachelor MoBi/Info/Bioinfoprojekt/NES.enrichment_matrix.RDS")
#saveRDS(C6.new.genesets, "/Users/paulaverdung/Desktop/Uni/Bachelor MoBi/Info/Bioinfoprojekt/C6.new.genesets.RDS")
```
I can also use the heatmap to identify clusters of patients
```{r}
k = 4
col_dendrogram_enrichment_matrix = heatm.patietns.gsea$tree_col
col_clusters = cutree(col_dendrogram_enrichment_matrix, k = k)

cluster_matrix = as.matrix(which(col_clusters == 1))
for (i in (2:k)){
  elements_clusters = which(col_clusters == i)
  cluster_matrix = cbind(cluster_matrix, elements_clusters)
}
colnames(cluster_matrix) = c(1:k) 
```
As the heatmap shows only poor clustering and some of the NES values are not even significant, we want to calculate the (adjusted) p values and set those NES values which are not significant to 0.
We decided to use the p values instead of the adjusted ones as the adjusted p values were substantially higher, probably due to around 15-20% of the values in the ranked expression data showing the same rank.
```{r}
pval.list.patients = list()
for (patient in names(named.numericals.patients)){
  stats.gsea.patients = named.numericals.patients[[patient]]
  gsea_patient_res = fgseaSimple(
    stats = stats.gsea.patients,
    nperm = 100,
    pathways = new_genesets
  )
  pval.list.patients[[patient]] = gsea_patient_res$pval
}
```
Now set according NES values to 0
```{r}
#store the adjusted p vals in a matrix of same format as the NES matrix
pval_matrix = as.matrix(pval.list.patients[[1]])
for (i in (2:58)){
  vec = as.matrix(pval.list.patients[[i]])
  pval_matrix = cbind(pval_matrix, vec)
}
#find positions in which p-values are above certain threshhold
pval.t = 0.15
nsig.NES = which(pval_matrix > pval.t)
adj.enrichment_matrix = enrichment_matrix
for (e in nsig.NES){
  adj.enrichment_matrix[e] = 0
}
```
```{r}
#saveRDS(adj.enrichment_matrix, "/Users/paulaverdung/Desktop/Uni/Bachelor MoBi/Info/Bioinfoprojekt/NES.enrichment_matrix.adj.RDS")
```
Plot the new heatmap
```{r}
adj.heatm.patietns.gsea = pheatmap(adj.enrichment_matrix, cluster_rows = , labels_row = names_new_genesets, fontsize_row = 5, labels_col = numbers_patients, fontsize = 7, color = colorRampPalette(brewer.pal(10, "RdBu"))(256)) 
```
Furthermore we want to compute the GSEA on the patients on the new gene sets we identified 
```{r}
NES.list.patients.C6 = list()
for (patient in names(named.numericals.patients)){
  stats.gsea.patients = named.numericals.patients[[patient]]
  gsea_patient_res = fgseaSimple(
    stats = stats.gsea.patients,
    nperm = 100,
    pathways = C6.new.genesets
  )
  NES.list.patients.C6[[patient]] = gsea_patient_res$NES
}
```
Enrichment Matrix:
```{r}
enrichment_matrix.C6 = as.matrix(NES.list.patients.C6[[1]])
for (i in (2:58)){
  vec = as.matrix(NES.list.patients.C6[[i]])
  enrichment_matrix.C6 = cbind(enrichment_matrix.C6, vec)
}

heatm.patietns.gsea.C6 = pheatmap(enrichment_matrix.C6, cluster_rows = , labels_row = names(C6.new.genesets), fontsize_row = 5, labels_col = numbers_patients, fontsize = 7) 
```
```{r}
#saveRDS(enrichment_matrix.C6, "/Users/paulaverdung/Desktop/Uni/Bachelor MoBi/Info/Bioinfoprojekt/NES.enrichment_matrix.C6.RDS")
```
Similarly to above we now want to remove the non-significant values of NES:
```{r}
pval.list.patients.C6 = list()
for (patient in names(named.numericals.patients)){
  stats.gsea.patients = named.numericals.patients[[patient]]
  gsea_patient_res = fgseaSimple(
    stats = stats.gsea.patients,
    nperm = 100,
    pathways = C6.new.genesets
  )
  pval.list.patients.C6[[patient]] = gsea_patient_res$pval
}
```
Pval Matix:
```{r}
#store the adjusted p vals in a matrix of same format as the NES matrix
pval_matrix.C6 = as.matrix(pval.list.patients.C6[[1]])
for (i in (2:58)){
  vec = as.matrix(pval.list.patients.C6[[i]])
  pval_matrix.C6 = cbind(pval_matrix.C6, vec)
}
#find positions in which p-values are above certain threshhold
pval.t.C6 = 0.1
nsig.NES.C6 = which(pval_matrix.C6 > pval.t.C6)
adj.enrichment_matrix.C6 = enrichment_matrix.C6
for (e in nsig.NES.C6){
  adj.enrichment_matrix.C6[e] = 0
}
```
```{r}
#saveRDS(adj.enrichment_matrix.C6, "/Users/paulaverdung/Desktop/Uni/Bachelor MoBi/Info/Bioinfoprojekt/NES.enrichment_matrix.C6.adj.RDS")
```
Plot the new heatmap:
```{r}
adj.heatm.patietns.gsea.C6 = pheatmap(adj.enrichment_matrix.C6, cluster_rows = , labels_row = names_new_genesets, fontsize_row = 5, labels_col = numbers_patients, fontsize = 7) 
```