LUSC = Lung squamous cell carcinoma

Filtering for caner_type_abbreviation == LUSC:
```{r}
df_LUSC = subset(tcga_annot, cancer_type_abbreviation == "LUSC")
```

Filtering for caner_type_abbreviation == LUAD:
```{r}
df_LUAD = subset(tcga_annot, cancer_type_abbreviation == "LUAD")
```

Creating two dataframes creating the expression values according to the selected sample IDs:
```{r}
install.packages("dplyr")
```

```{r}
library(dplyr)
# hier wähle ich aus dem pancancer dataset die jenigen spalten aus, bei denen die sample ID mit der sample ID aus den spezfischen LUSC und LUAD datasets (annot) übereinstimmen
sample_ids_LUSC = df_LUSC$sample
LUSC_exp = tcga_exp_proteins_shortened %>% select(intersect(colnames(tcga_exp_proteins_shortened), sample_ids_LUSC))

sample_ids_LUAD = df_LUAD$sample
LUAD_exp = tcga_exp_proteins_shortened %>% select(intersect(colnames(tcga_exp_proteins_shortened), sample_ids_LUAD))
```

```{r}
dim(LUSC_exp)
dim(LUAD_exp)
```


Creating the volcano plot following the same steps as in "DescriptiveAnalysisLUAD":

```{r}
# calculating the mean expressions for each gene in LUSC and LUAD tissue
LUSC_exp_means = apply(LUSC_exp, 1, mean)
LUAD_exp_means = apply(LUAD_exp, 1, mean)
```

Calculating the Foldchange value (LUAD/LUSC)
```{r}
LUSC_LUAD_FC = Map("-", LUAD_exp_means, LUSC_exp_means)
```

Now I am creating a dataframe with the gene ids and FC-values:
```{r}
df_FC_LUSC_LUAD = (do.call(cbind, LUSC_LUAD_FC))
df_FC_LUSC_LUAD = as.data.frame(t(df_FC_LUSC_LUAD))
```


```{r}
transposed_LUSC = t(LUSC_exp)
transposed_LUAD = t(LUAD_exp)
# musste man transponieren weil das über Spalten rechnet

# Vektor für die Speicherung der p-Werte
p_values_LL = numeric(ncol(transposed_LUSC))

for (i in 1:ncol(transposed_LUSC)) {
  tryCatch({
    shapiro_result_tumor = shapiro.test(transposed_tumor[, i])
    shapiro_result_normal = shapiro.test(transposed_normal[, i])
    
    p_values_LL[i] = min(shapiro_result_tumor$p.value, shapiro_result_normal$p.value)
  }, error = function(e) {
    p_values_LL[i] = NA  # Setze p-Wert auf NA, wenn ein Fehler auftritt
  })
}

# Anzeigen der p-Werte
#print(p_values4_LL

sumNALL = sum(is.na(p_values_LL))
print(sumNALL)

p_values_über_0.5_LUSC_LUAD = sum(p_values_LL > 0.05)
print(p_values_über_0.5_LUSC_LUAD)
print(length(p_values_LL))
ratio_not_normally_distributed_LL = p_values_über_0.5_LUSC_LUAD / length(p_values_LL)
print(ratio_not_normally_distributed_LL)
```
over 1/3 can not be considered to be normally distributed. Therefore, Wilcoxon-Test (unpaired) will be used to get the p values. 

```{r}
# Vektor für die Speicherung der p-Werte
p_values_LUSC_LUAD = numeric(ncol(transposed_LUSC))

for (i in 1:ncol(transposed_LUAD)) {
  wilcoxon_test_result_LL = wilcox.test(transposed_LUAD[,i], transposed_LUSC[,i], paired = F)
  
  p_values_LUSC_LUAD[i] = wilcoxon_test_result_LL$p.value
}
```

Bonferroni correction:
```{r}
bonferroni_alpha_LUSC_LUAD = 0.05 / length(p_values_LUSC_LUAD)
print(bonferroni_alpha_LUSC_LUAD)
```

adding the p_values in the dataframe:
```{r}
df_FC_LUSC_LUAD = cbind(df_FC_LUSC_LUAD, p_values_LUSC_LUAD)

df_FC_LUSC_LUAD = as.data.frame(df_FC_LUSC_LUAD)
```


Volcano plot:
```{r}
#install.packages(c("tidyverse", "ggrepel"))
#install.packages("kableExtra")

# Load packages
library(tidyverse)
library(ggrepel)
library(kableExtra)


# A short function for outputting the tables
knitr_table = function(x) {
  x %>% 
    knitr::kable(format = "html", digits = Inf, 
                 format.args = list(big.mark = ",")) %>%
    kableExtra::kable_styling(font_size = 15)
}

df_FC_LUSC_LUAD = df_FC_LUSC_LUAD %>% 
  mutate(
    Expression_in_LUAD = case_when(V1 >= 0 & p_values_LUSC_LUAD <= bonferroni_alpha_LUSC_LUAD ~ "Upregulated genes",
                           V1 <= 0 & p_values_LUSC_LUAD <= bonferroni_alpha_LUSC_LUAD ~ "Downregulated genes",
                           TRUE ~ "Unchanged genes")
    )
```

```{r}
library(ggplot2)
p_LUSC_LUAD = ggplot(df_FC_LUSC_LUAD, aes(V1, -log(p_values_LUSC_LUAD,10))) +
  geom_point(aes(color = Expression_in_LUAD), size = 2/5) +
  xlab(expression("log"[2]*"FC")) + 
  ylab(expression("-log"[10]*"p_value (adjusted)")) +
  scale_color_manual(values = c("dodgerblue", "gray50", "brown1")) +
  guides(colour = guide_legend(override.aes = list(size=1.5))) 
p_LUSC_LUAD
```
Using a boxplot to compare the differences between LUSC and LUAD for one gene (gene on the left):
```{r}
#install.packages("gridExtra")
library(gridExtra)
LUAD_LUSC_outlier = rownames(subset(df_FC_LUSC_LUAD, V1 < -10.6))
#print(LUAD_LUSC_outlier)

outlier_exp_LUAD = LUAD_exp[rownames(LUAD_exp) %in% LUAD_LUSC_outlier, ]
outlier_exp_LUSC = LUSC_exp[rownames(LUSC_exp) %in% LUAD_LUSC_outlier, ]

t_outlier_exp_LUAD = t(outlier_exp_LUAD)
t_outlier_exp_LUAD = as.data.frame(t_outlier_exp_LUAD)
t_outlier_exp_LUSC = t(outlier_exp_LUSC)
t_outlier_exp_LUSC = as.data.frame(t_outlier_exp_LUSC)

y_limits = range(c(t_outlier_exp_LUAD$ENSG00000134757, t_outlier_exp_LUSC$ENSG00000134757))

outlier_p1 = ggplot(t(outlier_exp_LUSC), aes(x = factor(0), y = ENSG00000134757)) +
  geom_boxplot() +
  labs(title = "LUSC", x = NULL, y = "ENSG00000134757") +
  theme_minimal() +
  ylim(y_limits)
outlier_p1

outlier_p2 = ggplot(t(outlier_exp_LUAD), aes(x = factor(0), y = ENSG00000134757)) +
  geom_boxplot() +
  labs(title = "LUAD", x = NULL, y = "ENSG00000134757") +
  theme_minimal() +
  ylim(y_limits)
outlier_p2

grid.arrange(outlier_p1, outlier_p2, ncol = 2)
```



```{r}
LUSC_LUAD_FC_values = unlist(LUSC_LUAD_FC)
LL_FC_sorted <- sort(LUSC_LUAD_FC_values)

bottom_1500 <- LL_FC_sorted[1:1500]
top_1500 <- tail(LL_FC_sorted, 1500)


#interesting_genes_LL_under4 = rownames(subset(df_FC_LUSC_LUAD, V1 < -4))

rownames_niedrig <- rownames(df_FC_LUSC_LUAD)[which(LUSC_LUAD_FC_values %in% bottom_1500)]

organism = "org.Hs.eg.db"

# Pathway-Analysis
pw_LL_under4 = enrichGO(gene = rownames_niedrig,
               OrgDb = organism,
               keyType = "ENSEMBL",
               ont = "BP",
               pvalueCutoff = 0.05)

dotplot(pw_LL_under4)
```

```{r}
#interesting_genes_LL_over4 = rownames(subset(df_FC_LUSC_LUAD, V1 > 4))

rownames_hohe <- rownames(df_FC_LUSC_LUAD)[which(LUSC_LUAD_FC_values %in% top_1500)]

organism = "org.Hs.eg.db"

# Pathway-Analysis
pw_LL_over4 = enrichGO(gene = rownames_hohe,
               OrgDb = organism,
               keyType = "ENSEMBL",
               ont = "BP",
               pvalueCutoff = 0.05)

dotplot(pw_LL_over4)
```


------------------------------------------------------------------
Now we compare LUAD / all other tumor types:

remove tumor type "LAUD" from tcga_exp:
```{r}
# ! = nicht
tcga_exp_withoutLUAD = tcga_exp_proteins_shortened[, !colnames(tcga_exp_proteins_shortened) %in% sample_ids_LUAD]
```

calculating means, FC and p values:
```{r}
# means
LUAD_exp_means = apply(LUAD_exp, 1, mean)
tcga_exp_withoutLUAD_means = apply(tcga_exp_withoutLUAD, 1, mean)

# FC
OTHER_LUAD_FC = Map("-", LUAD_exp_means, tcga_exp_withoutLUAD_means)

# creating a df
df_FC_OTHER_LUAD = (do.call(cbind, OTHER_LUAD_FC))
df_FC_OTHER_LUAD = as.data.frame(t(df_FC_OTHER_LUAD))

# transpose 
transposed_LUAD = t(LUAD_exp)
transposed_OTHER = t(tcga_exp_withoutLUAD)

# p values
p_values_OTHER_LUAD = numeric(ncol(transposed_OTHER))

for (i in 1:ncol(transposed_LUAD)) {
  wilcoxon_test_result_OTHER_LUAD = wilcox.test(transposed_LUAD[,i], transposed_OTHER[,i], paired = F)
  
  p_values_OTHER_LUAD[i] = wilcoxon_test_result_OTHER_LUAD$p.value
}

# bonferroni correction
bonferroni_alpha_OTHER_LUAD = 0.05 / length(p_values_OTHER_LUAD)
print(bonferroni_alpha_OTHER_LUAD)

# adding p values in df
df_FC_OTHER_LUAD = cbind(df_FC_OTHER_LUAD, p_values_OTHER_LUAD)
df_FC_OTHER_LUAD = as.data.frame(df_FC_OTHER_LUAD)
```

volcano plot:
```{r}
knitr_table = function(x) {
  x %>% 
    knitr::kable(format = "html", digits = Inf, 
                 format.args = list(big.mark = ",")) %>%
    kableExtra::kable_styling(font_size = 15)
}

df_FC_OTHER_LUAD = df_FC_OTHER_LUAD %>% 
  mutate(
    Expression = case_when(V1 >= 0 & p_values_OTHER_LUAD <= bonferroni_alpha_OTHER_LUAD ~ "Up-regulated",
                           V1 <= 0 & p_values_OTHER_LUAD <= bonferroni_alpha_OTHER_LUAD ~ "Down-regulated",
                           TRUE ~ "Unchanged")
    )

p_OTHER_LUAD = ggplot(df_FC_OTHER_LUAD, aes(V1, -log(p_values_OTHER_LUAD,10))) +
  geom_point(aes(color = Expression), size = 2/5) +
  xlab(expression("log"[2]*"FC")) + 
  ylab(expression("-log"[10]*"p_value")) +
  scale_color_manual(values = c("dodgerblue3", "gray50", "firebrick3")) +
  guides(colour = guide_legend(override.aes = list(size=1.5))) 
p_OTHER_LUAD
```
Using a boxplot to compare the differences between other tumor types and LUAD for one gene on the right:
```{r}
LUAD_OTHER_outlier = rownames(subset(df_FC_OTHER_LUAD, V1 > 14 & -log(p_values_OTHER_LUAD, 10) > 300))  
print(LUAD_OTHER_outlier)

outlier_exp_LUAD_O = LUAD_exp[rownames(LUAD_exp) %in% LUAD_OTHER_outlier, ]
outlier_exp_OTHER = tcga_exp_withoutLUAD[rownames(tcga_exp_withoutLUAD) %in% LUAD_OTHER_outlier, ]

t_outlier_exp_LUAD_O = t(outlier_exp_LUAD_O)
t_outlier_exp_LUAD_O = as.data.frame(t_outlier_exp_LUAD_O)
t_outlier_exp_OTHER = t(outlier_exp_OTHER)
t_outlier_exp_OTHER = as.data.frame(t_outlier_exp_OTHER)

y_limits_O = range(c(t_outlier_exp_LUAD_O$ENSG00000122852, t_outlier_exp_OTHER$ENSG00000122852))

outlier_p3 = ggplot(t(outlier_exp_OTHER), aes(x = factor(0), y = ENSG00000122852)) +
  geom_boxplot() +
  labs(title = "other cancer types", x = NULL, y = "ENSG00000122852") +
  theme_minimal() +
  ylim(y_limits_O)
outlier_p3

outlier_p4 = ggplot(t(outlier_exp_LUAD_O), aes(x = factor(0), y = ENSG00000122852)) +
  geom_boxplot() +
  labs(title = "LUAD", x = NULL, y = "ENSG00000122852") +
  theme_minimal() +
  ylim(y_limits_O)
outlier_p4

grid.arrange(outlier_p3, outlier_p4, ncol = 2)
```


```{r}
interesting_genes_OTHER_LUAD_over5 = rownames(subset(df_FC_OTHER_LUAD, V1 > 5))

organism = "org.Hs.eg.db"

# Pathway-Analysis
pw_OTHER_LUAD_over5 = enrichGO(gene = interesting_genes_OTHER_LUAD_over5,
               OrgDb = organism,
               keyType = "ENSEMBL",
               ont = "BP",
               pvalueCutoff = 0.05)

dotplot(pw_OTHER_LUAD_over5)
```

